\chapter*{Abstract}

Topological spaces express naturally in mathematical analysis, abstract algebra and geometry. This has made topology one of the great unifying ideas of mathematics. General topology, or point-set topology, defines and studies some useful properties of spaces and maps, such as connectedness, compactness and continuity.

\vskip .1cm

General topology is one of the thrust areas of topology in mathematics. A large number of mathematicians have contributed a lot in the development of general topology all over the world. They have used various types of sets to define and study various types of topological properties like covering and separation axioms. Side by side, they have generalized continuity and homeomorphism also according to the requirements. The primary goal in general topology is to investigate and compare different classes of topological spaces.

\vskip .1cm

Here the study is to focus on new type of closed sets namely ws-closed sets in topological spaces. In the growth of general topology, investigation of different types of generalized closed sets are of much interest for research scholars. We study the relationship between ws-closed sets and some existing closed sets in topological spaces. The basic properties of ws-closed sets and ws-open sets have been investigated. Using the concept of continuity, ws-continuous functions are introduced and compared with various existing continuous functions in topological spaces. The ws-closed maps, ws-open maps and ws- homeomorphisms in topological spaces are also introduced and investigated. Further, we study and investigate a new class of separation axioms namely ws-T k spaces (k = 0; 1; 2). Also, basic properties of these separation axioms are discussed. Using the concept of ws- open sets and ws-closed sets, ws-regular space and ws-normal spaces in topological spaces have been introduced and investigated. This study in topological space is also extended to ws-locally closed sets and ws-locally continuous functions. Finally, ws-closed sets and ws-continuous functions are studied in bitopological spaces.


