\chapter{{{\sl\bfseries Śāstra}\relax} -- an impediment to progress?}\label{chapter12}
\vskip -10pt

%ﬀ

\Authorline{Rajath V}
\lhead[\small\thepage\quad Rajath V]{}
\vskip -10pt

\section*{Abstract}

{\sl Śāstra} is of two types -- {\sl pauruṣeya} and {\sl apauruṣeya} -- based on, whether the subject matter revealed therein is accessible to be veriﬁed by the human mind or not; and whether the said {\sl śāstra} has an author (human) or not. Veda is {\sl apauruṣeya} since the subject matter it reveals, like {\sl dharmādharma-vyavasthā, punarjanma etc.}, can neither be proved nor disproved by the ﬁve senses and/or inference. Vedāṅga-s are ancillary sciences which help one in understanding the intended meaning of the Veda, while Upaveda-s, Itihāsa, Purāṇa, {\sl Smṛtigrantha}-s etc. are authored to percolate the vision of Veda to all sections of society and so on are considered {\sl pauruṣeya}.

	The contrarian view is that the Vedāṅga-s especially, Vyākaraṇa and Chandas which are meant to describe the grammar and prosody used in the Veda; somehow, over time became prescriptive to the extent that future users of the language were forced to adhere to them more strictly than done in the Veda itself. Such impositions can be found in the context of Dharmaśāstra-s, Āyurveda etc. In this way, it is claimed that the discourse on the application of {\sl śāstra}-s which was descriptive and ﬂexible in spirit and to begin with, over time has become prescriptive and rigid, thereby stiﬂing the contributions from the creative intelligence of all sections of society.

	On the one hand, we see that the Indian society has been vigorous all through, making original contributions in the ﬁeld of science and technology, mathematics, economics, philosophy and language; but on the other hand is said to have gone into decline for a thousand years. This contradiction begs an effort to understand Indian history and culture in a holistic manner, paying adequate heed to all the underlying aspects.

\section{Introduction}\label{art12-sec1}

A critical and honest examination of Pollock's paper on Śāstra-s (Pollock 1985) will motivate the reader to undertake a thorough study of Indian history and the unique place {\sl Śāstra}-s hold in the Indian intellectual tradition. The study of Indian history is made complicated by the presence of various kinds of history -- Marxist, Nationalist etc. In the context of the ﬁeld of Indology being dominated by Westerners, many of whom are not sympathetic to the Indic civilization, Sri Aurobindo remarks ``...a time must come when the Indian mind will shake off the darkness that has fallen upon it, cease to think or hold opinions at second and third hand and reassert its right to judge and enquire in a perfect freedom into the meaning of its own Scriptures....'' (Aurobindo 2000:130) In recent times, lot of new evidences are coming through which challenge the long held positions about Indian history; prominent among them being ---
\begin{itemize}
\item[$\bullet$] Study of World Economic history where India led the world GDP for at least 1500 years (Maddison 2001)

\item[$\bullet$] Inﬂuence of Indian mathematics on Europe (Raju 2007)

\item[$\bullet$] Archaeological evidence from Bronze-age Harappan civilization suggesting that it is at least 8000 year-old and not just 5500 (Sarkar et al. 2016)

\item[$\bullet$] Re-examination of evidence; relating to Sandrokottos and Devānāmpriya Priyadarśī, and Buddhist chronicles such as {\sl Dīpavaṃśa, Mahāvaṃśa}; pushing the date of the Buddha back by about 500 years (Arya 2015:286-292, Roy 2016)
\end{itemize}

Anybody who gets to know about the science \& technology, mathematics \& astronomy and economic history of ancient India and China seeks an explanation as to how and why the two countries paled into insigniﬁcance at the turn of the 20th century. In the Indian context, many scholars lay the blame for this decline at the doors of the invasions starting from Mahmud Ghazni to the advent of the British. This explanation is not entirely satisfying; for, prior to those events many invasions (Greek, Huns, Shakas etc.) were warded off and the refugees (Jews, Parsees etc.) were seamlessly absorbed into the Indian community (Sanyal 2008:5-19). Swami Vivekananda observes that with the overemphasis of {\sl nirvāṇa} for the masses in Buddhism \& Jainism, the society lost its {\sl kṣātra} and became vulnerable militarily to foreign invasions$^{1}$; and in turn led to an erosion of value systems with extraneous lifestyle, definitions of life goals and thought systems being imitated without proper examination. 

Sanyal argues that the decline in the Indian civilization was due to a change in mindset; described as an inward-looking tendency, not open to comparison and sharing knowledge with others. Its most glaring effect is that crossing the ocean became a taboo to the people who had a great ship-building industry and maritime expertise (Sanyal 2010:19-24)!

	In this paper, we closely examine another interesting idea proposed to explain this decline -- that {\sl śāstra}-s are an impediment to creativity and progress, forwarded by Pollock$^{2}$ . The rest of the paper is organized as follows: \$\ref{art12-sec2} (the next one) presents a traditional overview of {\sl śāstra}-s, \$\ref{art12-sec3} presents some of Pollock's views that {\sl śāstra}-s come in the way of progress and \$\ref{art12-sec4} concludes.

\section{{{\sl\bfseries Śāstra}\relax}-s --- Insiders' view}\label{art12-sec2}

The basic details of how the tradition views {\sl śāstra} will not be found explicitly discussed in the main source texts such as the Veda-s, the {\sl Bhagavad Gītā} or the Brahmasūtra-s. That is one of the reasons why it has led to so many concepts and interpretations different from the intention of {\sl śāstra}. Hence, it has been mandatory to learn the {\sl śāstra}-s from a guru who is grounded in the tradition, described as ``insider'' (Malhotra 2016:24-28). Some of the very important features of {\sl śāstra} are discussed in this section according to the tradition of Advaita Vedānta (since Pollock quotes {\sl Śaṅkarabhāṣya} in a few places), but I believe these details are acceptable to other Vedāntic traditions by and large.

\subsection{Definition \& Scope}\label{art12-sec2.1}

Etymologically, the word {\sl Śāstra} is derived in two ways ``{\sl śaṃsanāt śāstram}'' and ``{\sl śāsanāt śāstram}'', which indicate the {\sl pāramārthika} (transcendental) and {\sl vyāvahārika} (transactional) relevance of {\sl śāstra} respectively$^{3}$. {\sl Śāstra}, in the restricted sense of the term, refers to Veda (literally, `knowledge'), which primarily reveals those facts which cannot otherwise be known and Veda has no role in revealing what can be known by other available means (\$\ref{art12-sec2.2}). The subject matter of the Veda is the nature of {\sl jīva} (individual), reason for a particular birth, what happens after death, what means can be adopted while living to reach favorable ends after death etc. called {\sl dharmādharma-vyavasthā} (cosmic moral order), {\sl gati} (travel after death), {\sl loka} (other worlds) etc. Therefore, unless one has total {\sl śraddhā} in the words of the Veda, one cannot commit oneself wholeheartedly to understand and live this vision. 

The first part of the Veda, called Karmakāṇḍa reveals relation between  {\sl sādhana} and {\sl sādhya} (means \& ends) and is put in the form of injunctions and prohibitions ({\sl vidhi-niṣedha}). Knowledge of this part of Veda, called Aparā-vidyā (literally, ``lower'' `knowledge') serves two purposes -- one, to fulﬁll desires of man for a better body, better {\sl loka} \& possessions; while doing so the rituals demand 
\begin{itemize}
\item[$\bullet$] {\sl yajña} - performance of the sacred rites with a reverential attitude 

\item[$\bullet$] {\sl dāna} - sharing one's possessions 

\item[$\bullet$] {\sl tapas} - austerities$^{4}$ 
\end{itemize}

This brings about a tremendous change in the doer, committed to such a lifestyle to gain freedom from pride, arrogance, violent reactions, hoarding, selﬁshness \& pettiness etc., which in turn gives a mind, subtle enough for (or prepared to be open to) the second part of Veda called Jñānakāṇḍa or Vedānta, where there is only revelation of a fact already existing and self-evident. In this right seeing, called Parā-vidyā (literally, ``higher'' `knowledge'), the burden of do’s \& don't's have no more relevance. Without Aparā-vidyā, Parā-vidyā is impossible and without Parā-vidyā, Aparā-vidyā is incomplete. Use of the `faculty of choice' (unique to man) to gain and live this vision is the highest contribution one can make to oneself and to the entire creation at a very fundamental level.

\subsection{{{\sl\bfseries Pramāṇa-śāstra}\relax} or Epistemology}\label{art12-sec2.2}

That by which the existence of a thing is proved is called a {\sl pramāṇa}. In other words, any knowledge is valid only if it is born out of a {\sl pramāṇa}. The deﬁnition of a {\sl pramāṇa} is {\sl anadhigata, abādhita, arthavat bodhaka. Bodhaka} indicates the revealing nature of a {\sl pramāṇa}, with the other three words explained thus:

{{\sl\bfseries Anadhigata}\relax} -- that in the absence of a given {\sl pramāṇa}, one cannot even suspect ignorance of the corresponding object. For example, a blind person cannot even suspect his ignorance of color or form; nor does a person unexposed to Veda suspect his ignorance of {\sl dharmādharma-vyavasthā} and Brahman.

{{\sl\bfseries Abādhita}\relax} -- that one {\sl pramāṇa} can neither verify nor negate the knowledge acquired from another. For example, what the eye reveals (forms and colors) can neither be veriﬁed nor negated by the ears, nose etc. Similarly, what the Veda reveals can never be subjected to veriﬁcation by empirical sciences; which is based on data gathered through the ﬁve senses, or their extensions (telescope/microscope etc.), and inference based on such data. This is in contrast with modern science where a huge premium is placed on veriﬁcation.

{{\sl\bfseries Arthavat}\relax} -- that what is revealed by every {\sl pramāṇa} has some utility. 

The following six {\sl pramāṇa}-s are accepted to varying degrees in different {\sl darśana}-s in keeping with their dispositions and reasoning (Satprakashananda 2005):
\begin{itemize}
\item[(a)] {\sl Pratyakṣa} Direct Perception (via the five senses)

\item[(b)] {\sl Anumāna} Inference

\item[(c)] {\sl Upamāna} Example

\item[(d)] {\sl Arthāpatti} Two step inference (usually of the form `unless-otherwise')

\item[(e)] {\sl Anupalabdhi} Non-perception

\item[(f)] {\sl Śabda} Verbal testimony
\end{itemize}

In the cognition of any object, the instrument of knowledge is called {\sl pramāṇa}, the object is called {\sl prameya}, the cognizer is called {\sl pramātṛ}. When {\sl prameya} is not {\sl pratyakṣa}, it is either {\sl parokṣa} (remote) or {\sl aparokṣa} (of the nature of {\sl pramātṛ}). {\sl Parokṣa} itself can be either {\sl laukika parokṣa} (which can become {\sl pratyakṣa}) or {\sl nitya-parokṣa} (eternally remote, can never become {\sl pratyakṣa}) such as {\sl dharmādharma-vyavasthā, gati}, other {\sl loka}-s etc. (\$\ref{art12-sec2.1}). The role of {\sl śabda-pramāṇa} or Veda is in revealing {\sl nitya-parokṣa} and {\sl aparokṣa}.

When a person knows a bottle, he also knows that he has the knowledge of the bottle. How does he know this? Advaita-vedānta says this `knowledge' is {\sl svataḥ-siddha} (self-existent) and {\sl svataḥ-prakāśa} (self-eﬀulgent) which illumines all thoughts and objects (Satprakashananda 2005:110-112). Such a self-existent, self-eﬀulgent reality is called `Brahman' in the Upaniṣad-s -- ``{\sl prajñānaṃ brahma}'' ({\sl Aitareya Upaniṣad} 3.1.3), and is pointed out as the revealer ``{\sl na tatra sūryo bhāti ... tameva bhāntam anubhāti sarvaṃ tasya bhāsā sarvamidaṃ vibhāti}'' ({\sl Kaṭha Upaniṣad} 2.5.15). This Brahman obtaining as the Self is the real {\sl pramāṇa} (proof or evidence); since it alone reveals {\sl sukha-duḥkha} (pleasure and pain), the states of dream, sleep etc., while the {\sl pramāṇa}-s listed above are conﬁned to objects in the waking state only. Nonetheless, all the {\sl pramāṇa}-s indirectly intimate the Self alone--just as the cognition of all objects in a room intimate the presence of light--without which no knowledge can arise. It is for this reason that all branches of knowledge are revered in Indian culture, and are attributed to divine origin. Unless all knowledge born out of {\sl pramāṇa}-s does not resolve into the Self-evident, there is no end to seeking veriﬁcation.

\subsection{Vision about creation}\label{art12-sec2.3}

As the dictum of the Sāṅkhyas -- ``{\sl bhogāpavargārthau sṛṣṭiḥ}''$^{5}$ expounds -- ``creation'' is for the sake of jīva for the two-fold purpose of bhoga -- ({\sl karmakṣaya}/experiences of pain and pleasure) and {\sl apavarga} (emancipation). But at the same time, since no single jīva is responsible for the creation, there must be an intelligence which has given rise to this magniﬁcent, enormous and orderly creation; posited as Īśvara--therefore it is {\sl cetanakāryavāda} (since Īśvara cannot be inert) (``{\sl janmādyasya yataḥ}'' {\sl Brahma Sūtra} 1.1.2)$^{6}$. This is in stark contrast with modern science, where the purpose of creation is not known, and creation is said to have come about from matter. 

Secondly, just as how a pot cannot be brought into manifestation by a potter unless it is potential (unmanifest) in clay; similarly ``creation'' that is unmanifest alone is brought to manifestation. In other words, no new object is ever `created', and whatever comes into existence is that which is already in existence in an unmanifest, but potential, form. So, essentially ``creation'' is really a manifestation of names and forms while dissolution is unmanifestation and this cycle of creation and dissolution (manifestation and unmanifestation) proceeds according to the inviolable law of {\sl karma}. This view, which is the ultimate extension of the law of conservation, is called {\sl satkāryavāda} and is held by the Sāṅkhya and Vedānta {\sl darśana}-s.

\subsection{{{\sl\bfseries Śruti-Smṛti}\relax} demarcation}\label{art12-sec2.4}

Brahman, says the Upaniṣad, cannot be objectiﬁed by word or thought ``{\sl yato vāco nivartante aprāpya manasā saha ...}'' ({\sl Taittirīya Upaniṣad} 2.9.1), but is very much the reason for the functioning of the speech and mind ``... {\sl manaso mano yadvāco ha vācam} ...'' ({\sl Kena Upaniṣad} 1.2); and therefore is {\sl apauruṣeya-vastu}. That {\sl śāstra} which reveals this {\sl apauruṣeya-vastu} is called {\sl apauruṣeya-śāstra} the same as --{\sl Śruti} or Veda--in the tradition. It consists of the eternal truths discovered or ``seen'' by {\sl ṛṣi}-s ({\sl mantra-dṛṣṭṛ}-s) about {\sl dharmādharma-vyavasthā, gati, loka} etc. as discussed (\S\ref{art12-sec2.1}). Works authored by sages by applying the revelations of {\sl Śruti} to contemporary society are called {\sl Smṛti}-s. The {\sl Smṛti}-s are subservient to {\sl Śruti} since every author comes with his/her own interpretations and opinions. This distinction of {\sl Śruti}  and {\sl Smṛti}, of the separation the eternal principles and time-space dependent social norms, is unique to Hinduism; and it is this feature that enables Hinduism to constantly adapt to changing circumstances and be a living tradition continuously. Over time, unfortunately, this clarity of understanding is lost and as a consequence {\sl śraddhā}, the sacredness attached to this wisdom was lost; and blind, mechanical ritualistic beliefs stand to represent the most exalted tradition, giving place to contradictory opinions about {\sl śāstra}$^{7}$.

\subsection{{{\sl\bfseries Prakriyā}\relax} -- methodology}\label{art12-sec2.5}

\subsubsection{{{\sl\bfseries Adhyāropa-apavāda}\relax} (deliberate superimposition and subsequent negation)}\label{art12-sec2.5.1}

The subject matter of the Upaniṣad is so subtle that the Upaniṣad itself says that neither word nor thought can reveal (\$\ref{art12-sec2.4}), but yet it is possible to understand (``{\sl ānandaṃ Brahmaṇo vidvān}'', {\sl Taittirīya Upaniṣad} 2.9.1). This is not a contradiction but a paradox, which {\sl śāstra} itself resolves by resorting to a unique {\sl prakriyā} called {\sl adhyāropa-apavāda} (superimposition-negation) to communicate that which cannot be objectiﬁed by words ``{\sl adhyāropāpavādābhyāṃ niṣprapañcaṃ prapañcyate}''$^{8}$. At a superﬁcial glance, the statements belonging to the genres of {\sl adhyāropa} and {\sl apavāda} appear contradictory and might lead to confusion about the {\sl tātparya} or {\sl vivakṣā}, the intended meaning. Therefore, one has to carefully understand each statement in its context and also know the vision of the entire {\sl śāstra}, hence the need for a {\sl śrotriya-guru} (traditional master).

\subsubsection{Ontology -- {{\sl\bfseries Śruti, Yukti}\relax} and {{\sl\bfseries Anubhava}\relax}}

A marked diﬀerence between ontology in Indian traditions and that of the West is that, the Indian tradition takes into account all the three states of experience -- waking, dream \& sleep -- into consideration; whereas Western Science and philosophies examine only the waking state predominantly. Three realities are accepted in Advaita-vedānta viz.,
\begin{itemize}
\item[(a)] {{\sl\bfseries Prātibhāsika-satya}\relax}  available only for perception but lacks material existence -- like mirage-water, bending of stick partially immersed in water (due to refraction) etc.

\item[(b)] {{\sl\bfseries Vyāvahārika-satya}\relax} the material world that obtains in the waking state and is available for transaction.

\item[(c)] {{\sl\bfseries Pāramārthika-satya}\relax} that which is unnegatable in the three periods of time (past-present-future), and the ground in which waking, dream and sleep are all available.
\end{itemize}

The above three {\sl satya}-s with respect to a {\sl jīva} are his individuality, his body-mind-sense complex and Brahman (the essential nature) respectively. There are diﬀerences between {\sl darśana}-s on the nature and existence of {\sl pāramārthika-satya}, however all of them agree that the ``reality'' has to be in accordance with {\sl Śruti} (Veda), {\sl yukti} (logic) and {\sl anubhava} (experience). (``{\sl anubhavānusārī śrutiḥ.śrutyanusārī yuktiḥ}.'')

\subsubsection{{{\sl\bfseries Adhikaraṇa}\relax}-method}

To resolve apparent contradictions in any work where the author is not around, or for which there is no single author; the tradition has come up with the {\sl adhikaraṇa} method. In the case of the Veda-s, Jaimini and Bādarāyaṇa have composed their {\sl sūtra}-s keeping this methodology in mind which is adhered to by the commentators also; the salient features comprising the following:
\begin{itemize}
\item[(a)] {{\sl\bfseries Viṣaya}\relax} subject matter to be communicated

\item[(b)] {{\sl\bfseries Viśaya}\relax} possible doubts

\item[(c)] {{\sl\bfseries Pūrvapakṣa}\relax} presentation of an opposing viewpoint (or objection)

\item[(d)] {{\sl\bfseries Siddhānta}\relax} reconciliation \& presentation of the intended meaning of the {\sl śāstra}

\item[(e)] {{\sl\bfseries Saṅgati}\relax} connection between a topic and the next 
\end{itemize}

The feature that is most relevant in the context of this paper is {\sl pūrvapakṣa. Pūrvapakṣa} is an understanding or presentation of the opposing viewpoint objectively using its own terminology backed by its own reasoning. Once this is accomplished, the loopholes or limitations of pūrvapakṣa become clear in the light of the {\sl siddhanta} and a wrong understanding is resolved. For the {\sl pūrvapakṣa-siddhānta} dialogue to be fruitful, both parties must base their arguments on a methodology and a set of {\sl pramāṇa}-s (\$\ref{art12-sec2.2}) agreed upon by both. 

\subsubsection{{\sl\bfseries Anubandha-catuṣṭaya}}\label{art12-sec2.5.4}

{\sl Śāstra} is intended for the ultimate well-being ({\sl parama-hita}) of humans. The four-fold pre-requisites of every {\sl śāstra} are:
\begin{itemize}
\item[(a)] {{\sl\bfseries Prayojana}\relax} primary requirement on the part of {\sl śāstra} (and {\sl upadeśa}) is to point out the utility of the study at hand

\item[(b)] {{\sl\bfseries Adhikārin}\relax} qualiﬁcations of the seeker.

\item[(c)] {{\sl\bfseries Viṣaya}\relax} subject matter

\item[(d)] {{\sl\bfseries Sambandha}\relax} connection between {\sl viṣaya} and prayojana$^{9}$
\end{itemize}

It is very important to recognize the role of {\sl sambandha}--In one type of {\sl śāstra}, the knowledge gained has to be followed by action in order to achieve the desired end. For example, in the case of {\sl pākaśāstra} (treatise on cookery), after knowing a recipe, appropriate materials, utensils have to be procured and procedure followed till the dish is ready and ﬁnally has be eaten for satisfaction of hunger. The procedure to be followed ({\sl vidhi-niṣedha}) is a necessity for the desired end, and not to be looked upon as authoritative injunction. On the other hand, in the case of Vedānta, acquiring Self-knowledge alone is suﬃcient for {\sl mokṣa}, without needing anything else; since the end to be gained is not away from the seeker and the cause of his suﬀering was his mistaken identity. Therefore, Self-knowledge alone is the means and the end to see things as they are.  ``{\sl tarati śokam ātmavit}'' ({\sl Chāndogya Upaniṣad} 7.1.3)

In Vedānta, the qualiﬁcations of an {\sl adhikārin} are mentioned as {\sl mumukṣutva, vairāgya, viveka} and {\sl śama-dama-uparati-titikṣā-śraddhā-samādhāna}. {\sl Śraddhā} is often translated as `faith' or `belief'; but the correct translation in keeping with the spirit of the tradition is `openness'; an openness to the possibility of an intelligence not conﬁned to mind/person. Juluri presents this succinctly ``Hinduism to me is more than a religion, and even more than a way of life. To me, Hinduism is way of knowing life, intelligently.'' (Juluri 2015:6)

\section{Pollock's views}\label{art12-sec3}

Pollock is a prominent American Indologist who has infused new vigor into the ﬁeld of Indology which was losing its steam, by propounding a new outlook called `political philology'$^{10}$, according to which, the sacred texts of a civilization under study are examined from a political lens, where one set of people oppress another, pretty much in line with the famous `class-struggle', propounded by Karl Marx. Unlike most academicians who are satisﬁed with publishing papers and books, Pollock strives to bring about a positive change in society worldwide in general, and in marginalized sections of Indian society in particular; an exercise which he calls `liberating philology' (Pollock 2015).

Elsewhere in his literature, Pollock accepts that all narratives are equally valid (Naik 2016) --the insider's narrative is biased towards a sympathetic view of his civilization, while the outsider's narrative is biased towards a critical view of the same. He then argues that his bias is better since it releases the marginalized sections from oppression and sets them free.

\subsection{Pollock's Paper on {{\sl\bfseries Śāstra}\relax}}\label{art12-sec3.1}

Pollock starts on the note that ``{\sl śāstra} is one of the most fundamental features and problems of Indian civilization in general and of Indian intellectual history in particular'' (Pollock 1985:499) and goes on to say that he will examine the idea and nature of {\sl śāstra} which has not been the object of sustained Indological scrutiny. Theory, he says, has been held to precede and govern practice, which is diametrically opposite to that in the West. The implication has been the idea that all knowledge is pre-existent and therefore progress consists in a regressive re-appropriation of the past. The eternality of the Veda is used to naturalize and de-historicize cultural practices -- two components in a larger discourse of power (Pollock 1985:499).

\subsection{Naipaul's complaint about Indian art}\label{art12-sec3.2}

At the outset, Pollock observes a trend of strict {\sl śāstric} codiﬁcation across the entire cultural spectrum of India leading him to concur with V. S. Naipaul's complaint that the art of India was ``limited by the civilization, by an idea of the world in which men were born only to obey the rules.'' But this quote of Naipaul actually appears in his review of the book {\sl Imperial Mughal Painting} by Stuart Cary Welch, in the following passage:
\begin{myquote}
``...He has chosen forty pictures to illustrate the glory, limitations, and decline of Mughal court painting. An art that developed so fast, had Persia, India, and Europe to draw on...why didn't it do more? Why did this art, so human in the beginning, in the late sixteenth century, and so full of possibility, exhaust itself so quickly by the end of the seventeenth?

The answer can be inferred from Imperial Mughal Painting. The art was limited by the civilization, by an idea of the world in which men were born only to obey the rules--Islam was always in the wings, waiting to resimplify and stiﬂe. The art was limited by the despotism that went with this idea, the despotism that dealt only in power and glory but could create no nation. The art was limited by the ignorance and absurd conceit of a court dazzled by its own glitter. It shows in the paintings...'' (Naipaul 1979:10)
\end{myquote}

In this case, Pollock seems to be falling into the trap he is accusing the {\sl śāstra}-s of committing -- ignoring historicity, cultural change and socio-political conditions; thereby {\bf quoting out of context}. It is rather strange that Pollock does not ﬁnd any other observation of Naipaul worthy of quote, although Naipaul had reviewed at least four other books on Indian art in the very same article.

\subsection{Cultural grammars}\label{art12-sec3.3}

Pollock states that it is clearer to him that ``everywhere civilization as a whole...is constrained by rules of varying strictness, and indeed, may be accurately described by an accounting of such rules.'' (Pollock 1985:499) and goes on to compare the regulation of behavior between the {\sl Manusmṛti} and Amy Vanderbilt's {\sl Everyday Etiquette} in the case of one person meeting another. He further goes on to say that while in the West, such codes of etiquette have largely remained `tacit' knowledge existing on the level of practical and not discursive awareness; while in India, they were textualized at an early date and had ``to be learned rather than assimilated by a natural process of cultural osmosis.'' (Pollock 1985:500) 

While there may be some truth in the above analysis, it has more to do with a misunderstanding of {\sl śāstra} as pointed out in \$\ref{art12-sec2.4}. The text of the {\sl Manusmṛti} consists of four major divisions ---
\begin{itemize}
\item[(i)] {\sl sṛṣṭi-prakriyā} (origin of the world)

\item[(ii)] sources of {\sl dharma}

\item[(iii)] {\sl varṇāśrama-dharma} ({\sl dharma} of the social classes and stages of life)

\item[(iv)] {\sl karma-niyati} and {\sl mokṣa} (law of {\sl karma} and liberation).$^{11}$
\end{itemize}

The fourfold sources of {\sl dharma} are presented by the {\sl Manusmṛti} as -- Veda (or {\sl Śruti}), {\sl Smṛiti, sadācāra} (conduct of the noble people) and {\sl ātmanassantuṣṭi} (satisfaction of one's conscience) ({\sl Manusmṛti} 2.6 and 2.12). Pollock seems to suggest that because these {\sl dharmaśāstra}-s are influenced by the paradigm deriving from strict regulation of ritual action in Vedic ceremonies, these grammars were invested with massive authority ensuring a nearly unchallengeable claim to normative control of cultural practices. In fact, the view of {\sl Manusmṛti} itself is quite the opposite; that when social conditions change, the {\sl Manusmṛti} can be and has to be given up in favor of a better {\sl smṛti}. Justice M. Rama Jois states that
\begin{myquote}
``Yajnavalkya follows the same pattern as in Manu with slight modifications. On matters such as women's rights of inheritance and right to hold property, status of Sudras, and criminal penalty, Yajnavalkya is more liberal than Manu...'' (Jois 2004:31-32)
\end{myquote}
and suggests that the liberal nature of {\sl Yājñavalkyasmṛti} may have been influenced by Buddhism in ancient India; and perhaps that is the reason it includes prescriptions on the organization of monasteries, land grants, execution of deeds and other matters. (Jois 2004:31-32)

In the context of a {\sl vānaprasthī}, the {\sl Manusmṛti} states that he must gradually give up all attachments and being freed from opposites, remain in his true nature (Brahman) ({\sl Manusmṛti} 6.81); and ends on the note that one who sees the Self in all through understanding, he attains to the highest bliss, which has always been his true nature called Brahman ({\sl Manusmṛti} 12.125) (perfectly in line with the {\sl Upaniṣad}-s). Therefore, while the {\sl Manusmṛti} does prescribe detailed regulation in certain seemingly trivial matters (as Pollock presents the case of a meeting between two people); it is important to note that all the salient features of the Vedic vision are very well presented. It is a classic example, where the broad over-arching vision is kept in mind even while discussing the minutest detail.

\subsection{{{\sl\bfseries Śāstra}\relax}-s -- Definition \& Scope}\label{art12-sec3.4}

Pollock states that although the word {\sl śāstra} has occurred in ancient texts starting from the {\sl Ṛgveda}, no comprehensive deﬁnition is oﬀered till the medieval period! He then encounters two deﬁnitions of {\sl śāstra} -- ``authoritative rule'' and ``philosophical system''. Pollock seems to `struggle' to reconcile the above two deﬁnitions of śāstra and rather hastily concludes that the aspect of ``regulating'' or ``codifying'' is found to be prominent quoting Kumārilabhaṭṭa's   {\sl Ślokavārtika}. But this two-fold nature of {\sl śāstra} of revealing the ``vision'' and describing the ``means'' to realize it is very clear in the tradition (\$\ref{art12-sec2.1}) and is also succinctly captured in the {\sl Muṇḍakopaniṣad} and {\sl Bhagavadgītā} respectively as: 
\begin{myquote}
``{\sl parīkṣya lokān karmacitān brāhmaṇo nirvedamāyāt nāstyakṛtaḥ kṛtena} | 

{\sl tadvijñānārthaṃ...}'' ({\sl Muṇḍaka Upaniṣad} 1.2.12)

\vskip .1cm

``{\sl ārurukṣormuneryogaṃ karmakāraṇamucyate} |

{\sl yogārūḍhasya tasyaiva śamaḥ kāraṇamucyate} ||'' ({\sl Bhagavad Gītā} 6.3)
\end{myquote}

Then the {\sl vidyāsthāna}-s are presented by Pollock in line with Rājaśekhara's taxonomy, as follows ---

{\sl Apauruṣeya}
\begin{itemize}
\item[$\bullet$] 4 {\sl Vedas -- Ṛg, Yajus, Sāma, Atharva}

\item[$\bullet$] 6 {\sl Vedāṅga}-s -- {\sl Śikṣā, Kalpa, Vyākaraṇa, Jyautiṣa, Chandas, Nirukta}

\item[$\bullet$] 4 {\sl Upavedas} -- {\sl Āyurveda, Dhanurveda, Gāndharvaveda, Sthāpatyaveda}
\end{itemize}

Pauruṣeya
\begin{itemize}
\item[$\bullet$] {\sl Purāṇa and Itihāsa}

\item[$\bullet$] {\sl Ānvīkṣikī} -- logic \& philosophy

\item[$\bullet$] ({\sl Karma} and {\sl Brahma}) {\sl Mīmāṃsā}

\item[$\bullet$] {\sl Smṛtitantra} -- Dharmaśāstra-s
\end{itemize}

Others
\begin{itemize}
\item[$\bullet$] {\sl Vārttā} -- commerce

\item[$\bullet$] {\sl Daṇḍanīti} -- law

\item[$\bullet$] Erotology, Art, Architecture
\end{itemize}

In spite of noting that Rājaśekhara makes a distinction between {\sl apauruṣeya} and {\sl pauruṣeya śāstra}-s, Pollock again refers to the {\sl Ślokavārtika} ``{\sl śāstra} is that which teaches people what they should and should not do. It does this by means of eternal [words] or those made [by men]'' (Pollock 1985:501) and concludes that it undermines the dichotomy between {\sl pauruṣeya} and {\sl apauruṣeya} and claims that Madhusūdana Sarasvatī's {\sl Prasthānabheda} too abandons the said distinction. 

It is quite well known in the tradition that Pūrvamīmāṃsā gives importance to Vedic ceremonies as Jaimini puts it ``{\sl āmnāyasya kriyārthatvād-ānarthakyam-atadarthanām}'' ({\sl Jaiminī Sūtra} 1.2.1) (``The purpose of the Veda lying in the enjoining of actions, those parts of the Veda which do not serve that purpose are useless'' (Jha 1933:51)). It is therefore very clear that {\sl Ślokavārtika} won't pay much heed to the ``vision'' aspect of {\sl śāstra} since there is no place for injunctions there. It has been shown how these injunctions too are not really commandments, but a revelation of means for satisfying one's desires (\$\ref{art12-sec2.1}). While it is true that {\sl Prasthānabheda} does not explicitly mention {\sl pauruṣeya} and {\sl apauruṣeya}, a clear distinction is made between {\sl śāstra}-s which are authored and those which are not; with a list of authors and {\sl śāstra}-s composed by them. 

Pollock very simplistically translates {\sl pauruṣeya} and {\sl apauruṣeya} as human and transcendent respectively. The word ``transcendent'' does not capture the meaning of {\sl apauruṣeya}, since according to the Veda (especially {\sl Upaniṣad}-s), the {\sl apauruṣeya-vastu} called Brahman is both transcendent and immanent, and is not away from any particular experience--{\sl aparokṣa}. This is in stark contrast with the Semitic idea of a transcendent God completely removed from the world (\$\ref{art12-sec2.2}, \ref{art12-sec2.4}). Returning to the deﬁnition of {\sl śāstra} as ``philosophical system'', Pollock observes that there must be a possibility of `false {\sl śāstra}-s' or ``{\sl asacchāstra}-s'' and complains that while {\sl Manusmṛti} talks about it, Rājaśekhara's system cannot accommodate it. In addition, he says Mīmāṃsā too supports Rājaśekhara's system and invests {\sl śāstra}-s with inerrancy and paramount authority. (Pollock 1985:503)

The above statement of Mīmāṃsā supporting inerrancy of {\sl śāstra} is unfounded since Kumārila Bhaṭṭa's view on {\sl verbal} ({\sl scriptural}) {\sl cognition} is presented thus --
\begin{myquote}
“...the Vedas are not the work of a Personal Author; and being thus free from any defects due to such authorship, the Vedas must be regarded as the only source of knowledge (relating to {\sl Dharma}), which is infallible in its Self-Suﬃcient Validity.'' (Jha 1964:153)
\end{myquote}

Moreover, Rājaśekhara does not claim to propound a philosophical system; his intention is to present an analysis of literature and drawing far-reaching conclusions based his taxonomy is tantamount to putting words in his mouth. Therefore, the further claim that all {\sl śāstra}-s enjoy the unique qualities of inerrancy and paramount authority does not hold water.

\subsubsection{Change in character of {{\sl\bfseries Śāstra}\relax} from description to prescription}\label{art12-sec3.4.1}

Pollock observes that the oldest {\sl śāstra}-s, namely the Vedāṅga-s; especially Chandas (prosody), Vyākaraṇa (grammar) and Śikṣā (phonetics) started out as descriptive in character in helping to understand and assimilate the Vedic mantras. But, somehow over time, they changed their character to becoming prescriptive where future use of the language came to be governed by the above three {\sl śāstra}-s.

While it is not completely untrue that composers in Sanskrit adhered to Chandas, Vyākaraṇa and Śikṣā; it was largely because of the Sanskrit language being rich in vocabulary with hundreds of synonyms even for water, ﬁre etc. and a multitude of meters to choose from. Although it is argued that strict adherence to vyākaraṇa prevented the language from evolving freely, on the other hand Sanskrit works composed from the time of Pāṇini onward are perfectly intelligible today; unlike the vernaculars where, study of works about a thousand years old is not straightforward. In addition, grammar ﬁxed the framework of Sanskrit, but also provided for the creation of new words. In addition, grammar fixed the framework of Sanskrit, but also provided for the creation of new words. There are examples of Sanskrit words ({\sl gṛha}) that got into Prakrit ({\sl geha}) and were borrowed back into Sanskrit (Deshpande 1993:74). In pre-modern India, since majority of compositions were committed to memory, they were either in the form of cryptic pithy statements called {\sl sūtra}-s or metrical in nature called {\sl śloka}-s; and since knowledge was transferred from {\sl guru} to {\sl śiṣya}, any possible misunderstanding would be taken care of.

\subsection{Generalizing that theory is superior to practice}\label{art12-sec3.5}

Pollock quotes {\sl Chāndogya Upaniṣad} (``{\sl yadeva vidyayā karoti śraddhayā upaniṣadā tadeva vīryavattaraṃ bhavati}'' {\sl Chāndogya Upaniṣad} 1.1.10) that practice with {\sl vidyā} (knowledge of Om), {\sl śraddhā} (faith) and {\sl upaniṣad} is more eﬃcacious than mere practice without knowledge; and goes on to generalize that the intention of the {\sl śāstra}-s is to say that theory is superior to practice, citing examples from Vyākaraṇa, Āyurveda, {\sl Arthaśāstra} and {\sl Kāmasūtra}.
\begin{myquote}
The commentary on this mantra by Śaṅkarācārya goes thus --

``The Knowledge of the syllable `Om' being the essence of the `Essences', its being endowed with the qualities of fulﬁllment and prosperity does not consist merely in Knowledge of that syllable being a factor of the sacriﬁce: it is much more than that ... In ordinary life also, it is found that when a merchant and a forester sell pieces of Ruby and other gems, the former (who knows the real character of the gems) always obtains a higher price than the	latter (who is ignorant); and this is due to the superior Knowledge possessed by the merchant... The assertion that the Act of the man with Knowledge is `more eﬀective' than that of the ignorant man means that even when done by the ignorant person, the act is eﬀective; so that it does not mean that the ignorant man is not ﬁt to perform the act. In fact in the section dealing with {\sl Uṣasti} (later on) it is described that even ignorant persons have performed the priestly functions.

The act of meditating upon `Om' as the `essence of essences', as `fulfillment' and as `prosperity' forms a signle act of `meditation' (and worship); as there are no efforts intervening in between these...'' (Jha 1942:14-15)
\end{myquote}

The last statement forms the basis of the inclusiveness of `Hinduism', where all people are accepted as they are and an appropriate {\sl upāya} or {\sl sādhana} is presented for their gradual evolution. In a related context, the {\sl Bṛhadāraṇyaka Upaniṣad} states even more explicitly --
\begin{myquote}
``{\sl atha trayo vāva lokā manuṣyalokaḥ pitṛloko devaloka iti...karmaṇā pitṛloko vidyayā deva-loko deva-loko vai lokānāṃ śreṣṭhas tasmād vidyāṃ praśaṃsanti}.'' ({\sl Bṛhadāraṇyaka Upaniṣad} 1.5.16)
\end{myquote}
that the three {\sl loka}-s--{\sl manuṣyaloka, pitṛloka, devaloka}--are gained by {\sl putra} (begetting and taking care of progeny), {\sl karma} (ritualistic action) and {\sl vidyā} (meditation) respectively; and since {\sl devaloka } is the highest among the three, the means of attaining it viz. {\sl Vidyā} is praised. The meaning of the word {\sl vidyā} in this context is, therefore, to be taken as {\sl upāsana} or meditation, since mere knowledge cannot produce attainment of a {\sl loka} as its result.

Talking about the dialectical orientation prevalent in the medical tradition, Pollock says
\begin{myquote}
``... As is well known, the {\sl Suśrutasaṃhitā} urges the student to examine cadavers carefully and subject them to anatomical scrutiny, ``For it is only by combining both direct observation and the information of the {\sl śāstra} that thorough knowledge is obtained.'' The {\sl Carakasaṃhita} conﬁrms this view, ... proclaiming that ``The wise understand that their best teacher is the very world around them.''

Such voices ... may with some justice be viewed as oppositional, and in any case are pretty much in the minority. The dominant ideology is that which ascribes clear priority and absolute competence to shastric codiﬁcation...'' (Pollock 1985:510)
\end{myquote}

All {\sl darśana}-s accept {\sl pratyakṣa} as a valid {\sl pramāṇa} and hence saying that validity given to ``directly observed evidence'' is a minority or oppositional view is totally wrong. In fact, it  would be very interesting to examine the Buddhist and Jaina {\sl śāstra}-s since they do not accept {\sl śabda} as {\sl pramāṇa}; but this is conspicuously absent from this paper.

The word `theory' is etymologically derived from (Greek) {\sl theōros}, meaning spectator; but in English it has acquired the meaning of `a supposition or a system of ideas intended to explain something'. But, {\sl śāstra}, as discussed in \$\ref{art12-sec2.1} has a two-fold meaning -- one, revealing facts as they are; and the other, a manual for practical application. In other words, one part of {\sl śāstra} reveals the nature of existence as it is; and another part reveals the connection between a known action and a desired result$^{12}$. Based on the etymological derivation ``{\sl śāsanāt śāstram}'', {\sl śāstra} is said to be closer in meaning to `applied science'$^{13}$. Therefore, the language of theory and practice in the context of {\sl Śāstra} is too simplistic and in most occasions horribly wrong. 

In summary, the thesis that ``theory is superior to practice'' rests on lack of clarity in the meanings of words used, such as {\sl vidyā, śāstra}, theory etc. (or is quoted out of context); and the omission of a discussion on epistemology (the very basis for validity of {\sl śāstra}).

\subsection{Extrapolation of texts by {{\sl\bfseries Śiṣṭa}\relax}-s}\label{art12-sec3.6}

In the case of Dharmaśāstra-s, in a situation when the {\sl śruti} and {\sl smṛti} does not provide an explicit conclusion, one must look to {\sl śiṣṭācāra. Baudhāyana-dharmasūtra} deﬁnes the {\sl śiṣṭa}-s as devoid of jealousy, ego and other vices; and as well versed in {\sl dharma} who can extrapolate {\sl śāstra} to matters which it does not explicitly cover$^{14}$. Based on {\sl Śabara-bhāṣya} on the {\sl Jaimini-sūtra}-s -- ``{\sl dharmasya śabda-mūla-tvādaśabdam-anapekṣaṃ syāt}''  and ``{\sl api vā kartṛ-sāmānyāt-pramāṇamanu-mānaṃ syāt}'' ({\sl Jaiminī Sūtra} 1.3.1-2), Pollock states that ---
\begin{myquote}
``... the ``learned'' do not creatively reason from and extend {\sl śāstra} to illuminate obscure areas of social or moral conduct; on the contrary, their behavior derives directly from and fully conforms with the texts as codiﬁed, but these texts are ones to which we no longer have access.'' (Pollock 1985:506)
\end{myquote}

But, the {\sl Sūtra} and Śabara-bhāṣya actually state --
\begin{myquote}
[{\sl Sūtra} 1.3.1]: ({\sl Pūrvapakṣa}) ``Inasmuch as dharma is based upon the Veda, what is not Veda should be disregarded.'' (Jha 1933:87)

[{\sl Sūtra} 1.3.2]: ({\sl Siddhānta}) ``But ({\sl Smṛti}) is trustworthy, as there would be inference (assumption, of the basis in the Veda) from the fact of the agent being the same.'' 

[{\sl Bhāṣya}]: ``Even if they do not actually find it (a Vedic text as basis for the {\sl Smṛti} in question), they would infer it. It is quite possible also that the text upon which the {\sl Smṛti} is based was actually known to the {\sl Smṛti}-writer, but has since been forgotten.'' (Jha 1933:89)
\end{myquote}

The response given in this {\sl sūtra} and {\sl bhāṣya} to the objection presented in the previous {\sl sūtra} that `no non-vedic text ({\sl smṛti}) is valid,' is more in support of inference--in line with what Bodhāyana has said---than what is suggested by Pollock here. 

\subsection{Role of reasoning being condemned in {{\sl\bfseries Śāstra}\relax}}\label{art12-sec3.7}

Pollock quotes from the {\sl Bhagavadgītā} and Abhinavagupta's {\sl saṅgraha-śloka} as ---
\begin{myquote}
``Whoever abandons the injunctive rules of {\sl śāstra} and proceeds according to his own will never achieve success, or happiness, or ﬁnal beatitude...” (Pollock 1985:510)

``...one must never contemplate action according to one's own lights, but must instead follow shastric injunction...'' (Pollock 1985:510)
\end{myquote}
with the original {\sl śloka}-s being ---
\begin{myquote}
``{\sl yaḥ śāstravidhim utsṛjya vartate kāmakārataḥ} | {\sl na sa siddhimavāpnoti na sukhaṃ na parāṃ gatim} || {\sl tasmāc chāstraṃ pramāṇaṃ te kāryākāryavyavasthitau} |'' (Bhagavad Gītā 16.23-24)

\vskip .1cm
``{\sl abodhe svātma-buddhyaiva kāryaṃ naiva vicārayet} |

{\sl kintu śāstrokta-vidhinā śāstraṃ bodha-vivardhanam} ||'' (Raina 1933:162)
\end{myquote}

The above {\sl Gītā śloka} is said in the context where; under the spell of ignorance, when one is overcome by {\sl kāma} (passion), {\sl krodha} (anger) and {\sl lobha} (greed) -- beautifully summed up by Abhinavagupta in one word, {\sl abodha} (non-understanding) -- {\sl śāstra} must be the guiding light in determining what to do and what not to do. Here is a case of {\bf quoting out of context} by Pollock to suit his `theory'.

\subsection{Implications of Pre-existence of theory}\label{art12-sec3.8}

\subsubsection{Creation of knowledge as divine activity}\label{art12-sec3.8.1}

%raghu, page 19 (word)


