\chapter{anubaMdhagaLu} 

\section*{BAratiVya gaNitajacnxru}
\addcontentsline{toc}{section}{{\protect\bf BAratiVya gaNitajacnxru}}

{\renewcommand\labelitemi{}
\begin{itemize}
\item AyaRBaTa - {\rm 1} kirx.~sha. {\rm 476}
\item barxhamxgupatx {\rm 6 - 7} neya shatamAna
\item varAhamihirAcAyaR {\rm 6} neya shatamAna
\item (oMdaneya) BAsakxra {\rm 7} neya shatamAna
\item virahaMka kirx.~sha. {\rm 600 - 800}
\item viVraseVna kirx.~sha. {\rm 710 - 790}
\item sakxMdaseVna {\rm 8} neya shatamAna
\item padamxnABa
\item goVviMda sAvxminf kirx.~sha. {\rm 800 - 850}
\item lalalx {\rm 8} neya shatamAna
\item vararuci
\item shirxVdharAcAyaR {\rm 8} neya shatamAna
\item paqthUdakasAvxmi kirx.~sha. {\rm 850}
\item vaTeVshavxra kirx.~sha. {\rm 880}
\item mahAviVrAcAyaR {\rm 9} neya shatamAna
\item maMjuLAcAyaR {\rm 10} neya shatamAna
\item (eraDaneya) AyaRBaTa kirx.~sha. {\rm 950}
\item  jayadeVva kirx.~sha. {\rm 10} neya shatamAna
\item neVmicaMdarx {\rm 10} neya shatamAna
\item shirxVpati kirx.~sha. {\rm 1019 - 1066}
\item (eraDaneya) BAsakxrAcAyaR {\rm 12} neya shatamAna
\item halAyudha {\rm 11} neya shatamAna
\item rAjAdhitayx sumAru kirx.~sha. {\rm 1120}
\item savAyi jayasiMha (eraDaneya jayasiMha)
\item kamalAkAra
\item Thakukx raPeru {\rm 13} neya shatamAna
\item mAdhava kirx.~sha. {\rm 1350 - 1425}
\item parameVshavxra kirx.~sha. {\rm 1360 - 1455}
\item niVlakaMTha soVmayAji kirx.~sha. {\rm 1444 - 1544}
\item nArAyaNa paMDita {\rm 14} neya shatamAna
\item nArAyaNa mahiSa maMgalaM kirx.~sha. {\rm 1540 - 1610}
\item shaMkaravAriyarf {\rm 16} neya shatamAna
\item gaNeVsha deYvajacnx {\rm 16} neya shatamAna
\item acucxta piSAraTi kirx.~sha. {\rm 1550 - 1621}
\item jAcnxnarAja {\rm 15-16} neya shatamAna
\item sUyaRdAsa {\rm 16} neya shatamAna
\item muniVshavxra kirx.~sha. {\rm 1652 - 1718}
\item deYvajacnxvalalxBa {\rm 17 - 18} shatamAna
\item parxtamana soVmayAji kirx.~sha. {\rm 1660 - 1740}
\item shaMkara vamaRnf kirx.~sha {\rm 1800 - 1838}
\item shirxV rAmacaMdarx lAlf kirx.~sha. {\rm 1821 - 1880}
\item paMDita bApudeVvashAsitx {\rm 1821 - 1900}
\item ciMtAmaNi raGunAthA cAyaR {\rm 1828 - 1880}
\item sAmaMta caMdarxsheVKara siMha {\rm 1835 - 1904}
\item paM sudhAkara divxveVdinf {\rm (1855 - 1910)}
\item sarf. AshutoVSf muKajiR {\rm (1864 - 1924)}
\item shirxVrAmasAvxmi ayayxrf {\rm (1871 - 1936)}
\item sAvxmi rAmatiVthaRru {\rm (1873 - 1906)}
\item porx. parxboVdha caMdarx seVnf gupAtx {\rm (1876 - 1962)}
\item shirxV gaNeVsha parxsAda {\rm (1876 - 1935)}
\item porx biBUti BUSaNadatAtx {\rm (1888 - 1958)}
\item shirxVnivAsa rAmAnujanf kirx.~sha {\rm 1887 - 1920}
\item porx. e. kaqSaNxsAvxmi ayayxMgArf {\rm (1892 - 1953)}
\item porx. e. narasiMgarAvf {\rm (1893 - 1967)}
\item porx| parxshAMta caMdarx mahAlAnoVbisf {\rm (1893 - 1972)}
\item porx.~ke. AnaMdarAvf {\rm 1893 - 1966}
\item sateyxVMdarxnAtha boVsf kirx.~sha {\rm 1894 - 1974}
\item DA. Arf.veYdayxnAthasAvxmi {\rm (1894 - 1960)}
\item porx. ke.esf.ke aiyayxMgArf {\rm (1899 - 1944)}
\item porx. bi.enf parxsAda {\rm (1899 - 1966)}
\item porx. Ti.esf.kupapxNaNxshAsitxrX {\rm (1900 - 1982)}
\item DA. si.enf shirxVnivAsa ayayxMgArf {\rm (1901 - 1972)}
\item porx. esf.esf piLeLxY {\rm (1901 - 1950)}
\item DA. enf. siMgf {\rm (1901 - 1954)}
\item porx. Ti. vijayarAGuvanf {\rm (1902 - 1955)}
\item si.~Ti. rAjagoVpAlf {\rm (1903 - 1978)}
\item Di. Arf. kaperxVkarf {\rm (1905 - 1986)}
\item Di.~Di. kosAMbi {\rm (1907 - 1966)}
\item viSuNx jayaMta nAraLiVkarf {\rm (1908 - 1991)}
\item porx. ke. veMkaTAcalashAsitxrX {\rm (1908 - 2003)}
\item esf. caMdarxsheVKarf {\rm (1910 - 1995)}
\item Arf.~si. boVsf {\rm 20} neya 
\item porx. vasaMta shaMkara hujUrf bajArf
\item porx. si. rAdhAkaqSaNxrAvf {\rm 20} neya
\item jagadugxru sAvxmi shirxV BAratiV kaqSaNx tiVthaRru {\rm 20} neya
\item porx. ke. caMdarxsheVKarf {\rm (1920 - )}
\item DA. ke.~esf. nAgarAjanf
\item tala kulatAtxrf ra naMbUdiri
\item selalxlUra naMbUdiri
\item porx. pi.~elf. BaTAnxgarf {\rm (1912 - 1976)}
\item porx. miVnAkiSx suMdaraM {\rm (1913 - 1968)}
\item DA. Ti.~e. sarasapxtiVamamx {\rm (1918 - 2000)}
\item porx. ke.~esf. shukalx {\rm (1918 - 2007)}
\item porx. ke.~vi. shamaR {\rm (1919 - 2005)}
\item porx. ke.~ji.rAmanAthanf {\rm (1920 - 1992)}
\item DA hariVsha caMdarx {\rm (1923 - 1983)}
\item jagatf nArAyaNa kapUrf {\rm (1923 - 1983)}
\item porx. kaqSaNx Di. aBayxMkarf {\rm (1928 - 2007)}
\item porx. ke. rAmacaMdarx {\rm (1933 - 2011)}
\end{itemize}}

\newpage
\section*{pAshAcxtayx gaNitajacnxru}
\addcontentsline{toc}{section}{{\protect\bf pAshAcxtayx gaNitajacnxru}}

{\renewcommand\labelitemi{}
\begin{itemize}
\item apaloVniyasf kirx. pU. {\rm 260 - 200}, girxVsf
\item amAliV emimx noVyidarf jamaRni {\rm (1882 - 1935)}
\item arisATxTalf kirx. pU. {\rm 384 - 322}, girxVsf
\item AkiRmiDiVsf kirx. pU. {\rm 287 - 212}, girxVsf
\item AthaRrf keyilx kirx.~sha. {\rm 1821 - 1895}
\item AnAnxsATxPaDfR heMDfrikfsx {\rm (1905 - 2001)} yu.~esf.~e
\item Alf baTaTxni kirx.~sha. {\rm 877 - 929}, areVbiyA
\item Ayalxrf liyoVnADfR kirx.~sha. {\rm 1707 - 1783} sivxTajxrf lAyxMDf
\item Alf sivxnf kirx.~sha. {\rm 735 - 804}, iMgelxMDf
\item Alf kaKiR {\rm 11} neya shatamAna, areVbiyA
\item Alf biruni kirx.~sha. {\rm 973 - 1048} madhayx ESayx
\item Alf mamUnf kirx.~sha. {\rm 809 - 833}, areVbiyA
\item abelf kirx.~sha. {\rm 1802 - 1829}, nAveR
\item iveMjalisATx Torisililx kirx.~sha. {\rm 1608 - 1647} iTali
\item enAkisx mAMDalf kirx. pU. {\rm 611 - 547} girxVsf
\item emili DUyx cATele {\rm (1706 - 1749)} pAyxrisf
\item eraTAsfniVsf kirx.~pU. {\rm 274 - 194}, girxVsf
\item elijabetf bukananf kwle {\rm (1874 - 1945)} amerika
\item eliVnAlukerxjiyA kAnARro pisokxVpiyA {\rm (1646 - 1684)} venisfnagara
\item elekisxsf kADf kelxyaro kirx.~sha. {\rm 1713 - 1765} PArxnfsx
\item Omarf KayAM {\rm 12} neya shatamAna, areVbiyA
\item kADARnf kirx.~sha. {\rm 1501 - 1576}, iTali
\item kAyxMTarf kirx.~sha. {\rm 1855 - 1918}, jamaRni
\item kirxshacxna geVlfDxbeshf kirx.~sha. {\rm 1690-1764}
\item kumamxra kirx.~sha. {\rm 1789 - 1893}, jamaRni
\item kAlxrA lATimarf beVkanf {\rm (1866 - 1948)} nUyx iMgelxMDf
\item koVpaniRkasf kirx.~sha. {\rm 1473 - 1543}, poVleMDf
\item kwshxSi AgasiTxrXVnaM lUyi kirx.~sha. {\rm 1789 - 1857} PArxnfsx
\item korxnekarf kirx.~sha. {\rm 1823 - 1891}, jamaRni
\item Kinf kiyu SAvf {\rm 13} neya shatamAna, ciVna
\item gADf PerxY hAlfDx hADiR kirx.~sha. {\rm 1877 - 1947}, birxTanf
\item girADfR kirx.~sha. {\rm 1595 - 1632}, hAleMDf
\item gisepi piyojisx kirx.~sha. {\rm 1746 - 1829}, iTali
\item gAyxlAvx gelAyf kirx.~sha. {\rm 1811 - 1832} PArxnfsx
\item gwsf kAlfR PeDirxkf kirx.~sha. {\rm 1777 - 1855}, jamaRni
\item cebishevf kirx.~sha. {\rm 1821 - 1894} raSAyx
\item cAlfsx bAyxbeVjf kirx.~sha. {\rm 1792 - 1871} iMgelxMDf
\item jAnf neVpiyarf kirx.~sha. {\rm 1550 - 1617}, sAkxTfleMDf
\item jAnf vAlisf kirx.~sha. {\rm 1616 - 1703}, iMgelxMDf
\item jAnf vilasxnf kirx.~sha. {\rm 1741 - 1793}, iMgelxMDf
\item jAnfvenf kirx.~sha. {\rm 1834 - 1883}, iMgelxMDf
\item jAnf banURli kirx.~sha. {\rm 1667 - 1748}, sivxTasxrfleMDf
\item jiVnf joVsePf liveVriyarf kirx.~sha. {\rm 1811 - 1877} PArxnfsx
\item jeVmfsxgerxgori jiVnf kirx.~sha. {\rm 1638 - 1675}, sAkxTfleMDf
\item jeVmfsx banARli kirx.~sha. {\rm 1654 - 1705}, sivxTasxrfleMDf
\item joVhAnf kepalxrf kirx.~sha. {\rm 1571 - 1630}, jamaRni
\item JiVnoV (siVnoV) kirx. pU. {\rm 494 - 435}, girxVsf
\item TAlami kirx.~sha. {\rm 85 - 165} girxVsf
\item DalAM baTfR kirx.~sha. {\rm 1717 - 1783}, PArxnfsx
\item DayoVPAyxMTasf kirx.~sha. {\rm 250 - 275}, girxVsf
\item DimArxvf kirx.~sha. {\rm 1667 - 1754}, iMgelxMDf, (huTiTxdudx PArxnfsxnalilx)
\item DimAgaRnf kirx.~sha. {\rm 1806 - 1871}
\item DesAgURsf kirx.~sha. {\rm 1593 - 1662} PArxnfsx
\item DeviDf hilf baDfR kirx.~sha. {\rm 1862 - 1943}, jamaRni
\item thiyAnoV kirx. pU. {\rm 6} neV girxVsf
\item theVlfsx (thAlisf) kirx. pU. {\rm 640 - 556}, girxVsf
\item nikoVmAkasf (oMdaneya shatamAnada koneyalilx) girxVsf
\item nikoVlf oresimx kirx.~sha. {\rm 1323 - 1382}, PArxnfsx
\item niVnA kAloRvAnx bAri {\rm (1901 - 1961)} raSAyx
\item pApapxsf, girxVsf
\item pAnfselx kirx.~sha. {\rm 1788 - 1868}, PArxnfsx
\item pArxkalxsf kirx.~sha. {\rm 412 - 485}, girxVsf
\item pelxToV kirx. pU. {\rm 430 - 349}, girxVsf
\item peYthAgorasf kirx. pU. {\rm 572 - 501} girxVsf
\item PamAR kirx.~sha. {\rm 1601 - 1665}, PArxnfsx
\item PiboVnAkili, yoVnADoV kirx.~sha. {\rm 1170 - 1250} iTali
\item PUriyarf kirx.~sha. {\rm 1768 - 1830}, PArxnfsx
\item belxVsf pAsakxlf kirx.~sha. {\rm 1623 - 1662}, PArxnfsx
\item boVtiyasf kirx.~sha. {\rm 475 - 524} roVmf
\item maseVRnf kirx.~sha. {\rm 1588 - 1648}, PArxnfsx
\item mAnfSe kirx.~sha. {\rm 1746 - 1818}, PArxnfsx
\item mahamadf ibfnf mUsA {\rm 9} neya shatamAna, irAkf
\item menelAsf {\rm 1} neV shatamAna, girxVsf
\item menekamxsf kirx. pU. {\rm 375 - 325}, girxVsf 
\item meVriPeVrf PAyxkfsx soVmarf vile {\rm (1780 - 1872)} ATfkxleMDf
\item meVriya agenxVsi {\rm (1718 - 1799)} iTali
\item moVbiyasf AgasfTx PaDiRneMDf kirx.~sha. {\rm 1790 - 1868}, jamaRni
\item mAyxkisxmasf pAlxnUDfsx {\rm 14} neV shatamAna, TakiR
\item yUkilxDf kirx.~sha. {\rm 330 - 275}, girxVsf
\item yUDokasxsf kirx. pU. {\rm 408 - 355}, girxVsf
\item yoVhanf kepalxrf kirx.~sha. {\rm 1571 - 1630} jamaRni
\item rane DekATaR kirx.~sha. {\rm 1596 - 1650}, PArxnfsx
\item rabibxbena esarx kirx.~sha. {\rm 1093 - 1167}
\item lAgArxMjf josePf lUyisf kirx.~sha. {\rm 1736 - 1813} iTali
\item lApAlxsf perirx seYmanf kirx.~sha. {\rm 1749 - 1827}, PArxnfsx
\item lAbAceVSf simxV kirx.~sha. {\rm 1793 - 1856}, raSAyx
\item lAvo jenevarx seYmanfsx {\rm (1870 - 1949)} a. sa. saMsAthxna
\item lUyi keroVlf kirx.~sha. {\rm 1832 - 1898}
\item lUyi ekabxlf borxVgilx kirx.~sha. {\rm 1892 - 1960}
\item lejAMDfrx kirx.~sha. {\rm 1752 - 1833} PArxnfsx
\item leYsAMgf yaMgf {\rm (1952 - )} hAMkAMgf 
\item leYbinxjf kirx.~sha. {\rm 1646 - 1716} jamaRni
\item viliyaMPAyxMkfsx {\rm 10} neya shatamAna
\item viVTA {\rm 1540 - 1603} PArxnfsx
\item sarf AthaRrf eDiMgfTanf kirx.~sha. {\rm 1882 - 1941}
\item sarf aisAkf nUyxTanf kirx.~sha. {\rm 1642 - 1727} iMgelxMDf
\item sAMDarf sanf kirx.~sha. {\rm 1682 - 1739}, iMgelxMDf
\item siju e vU {\rm (1964 - )} ciVnA
\item seVkikoVva kirx.~sha. {\rm 1682 - 1708}, japAnf
\item soVPiV jamaRnf {\rm (1776 - 1831)} pAyxrisf
\item seTxYnarf kirx.~sha. {\rm 1796 - 1863}, sivxTasxrfleMDf
\item soVnAyx kovalevfsikx {\rm (1850 - 1891)} raSAyx
\item hamaRnf miMkosimx kirx.~sha. {\rm 1864 - 1909}
\item hAyxmilfTanf kirx.~sha. {\rm 1805 - 1865}, airfleMDf
\item hipApxkfRsf kirx. pU. {\rm 180 - 152}, girxVsf
\item hipApxkerxTiVsf kirx. pU. {\rm 470}, girxVsf
\item hilADx giVriMjarf PAnf miVsasf {\rm (1893 - 1973)} AsiTxrXya
\item henirx pAnfkAre kirx.~sha. {\rm 1854 - 1912} PArxnfsx
\item herAnf (hiVroV) {\rm 1} athavA {\rm 3} neV shatamAnada, girxVsf
\item henirx birxgfsx kirx.~sha. {\rm 1561 - 1631}, iMgelxMDf
\item heYginfsx kirxshicxyanf kirx.~sha. {\rm 1629 - 1695} nedarfleMDf
\item heYPATiyA kirx.~sha. {\rm 370 - 415}, girxVsf
\item hAyflx PerxDf {\rm 20} neya shatamAna iMgelxMDf
\end{itemize}

\newpage
\section*{jagadugxru BAratiVkaqSaNx tiVthaRru niVDiruva veVdagaNitada sUtarxgaLu}
\addcontentsline{toc}{section}{{\protect\bf jagadugxru BAratiVkaqSaNx tiVthaRru niVDiruva veVdagaNitada sUtarxgaLu}}

\begin{enumerate}[\rm 1)]
\item EkAdhikeVna pUveVRNa

\item niKilaM navatashacxramaM dashataH

\item UdhavxR tiyaRgABxyXmf 

\item parAvatayxR yoVjayeVtf 

\item shUnayxM sAmayx samucacxyeV

\item (AnurUpeyxV) shUnayxmanayxtf

\item saMkalana vayxvakalanABAyxmf

\item pUraNApUraNABAyxmf

\item calanakalanABAyxmf

\item yAvadUnamf

\item vayxSiTxsamaSiTxH

\item sheVSANayxMkeVna carameVNa

\item soVpAMtayxdavxyamaMtayxmf

\item EkanUyxneVna pUveVRNa

\item guNitasamucacxyaH

\item guNakasamucacxyaH
\end{enumerate}

\section*{pAriBASika(shabadxkoVsha) padagaLu}
\addcontentsline{toc}{section}{{\protect\bf pAriBASika(shabadxkoVsha) padagaLu}}

\begin{longtable}{>{\rm}l@{\hspace{1cm}}l}
Addition & kUDu\\
Algebra & biVjagaNita\\
Analogy & anurUpate, sAdaqshayx\\
Analyse & vishelxVSaNe\\
Application & anavxya\\
Area & sale\\
Arithmetic & aMkagaNita\\
Arithmetic Series & samAMtara sherxVNi\\
Artificial satellite & kaqtaka upagarxha\\
Ascending order & AroVhaNakarxma\\
Astrology & joyxVtiSayx\\
Astronomy & KagoVLa vijAcnxna, joVtiviRjAcnxna\\
Base number & AdhArasaMKeyx\\
Basic concept & mUla parikalapxne\\
Binomial theorem & divxpada parxmeVya\\
Brackets &  AvaraNagaLu\\
Calculus & kalanashAsatxrX\\
Circle & vaqtatx\\
Circumference & vaqtatxparidhi\\
Classification & vagiVRkaraNa\\
Code & saMkeVta\\
Combination & vikalapx\\
Commercial Mathematics & vANijayxgaNita\\
Complex number & saMkiVNaRsaMKeyx\\
Composite number & BAjayxsaMKeyx\\
Conclusion & tiVmARna\\
Conditional equation & nibaMdhita samiVkaraNa\\
Consecutive prime numbers & anukarxma aviBAjayx saMKeyxgaLu\\
Constant & sithxrAMka, niyatAMka\\
Counting numbers & eNike saMKeyxgaLu\\
Cube & Gana\\
Cuberoot & GanamUla\\
Cyclic Quadrilateral & cakirxVya catuBuRja\\
Decimal number & dashamAMsha saMKeyx\\
Decimal system & dashamAMsha padadhxti\\
Deductive & nigamana\\
Definition & vAyxKeyx\\
Denominator & CeVda\\
Descending & avaroVhaNa\\
Diameter & vAyxsa\\
Diagonal & kaNaR\\
Divider & viBAjaka\\
Divisibality & BAjayxte\\
Division & BAgAkAra\\
Divisor & BAjaka\\
Dodecahedron & karxmadAvxdashaPalaka; dAvxdasha\\[-0.1cm]
& muKa Gana\\
Equation & samiVkaraNa\\
Equilateral Triangle & samabAhutirxBuja\\
Estimate & aMdAju\\
Even number & samasaMKeyx\\
Euclidean geometry & yUkilxDiVya jAyxmiti\\
Factor & apavataRna\\
Fallacy & viroVdhABAsa\\
Febonacci numbers & PibonAkikx saMKeyxgaLu\\
Formula & sUtarx\\
Fraction & BinanxrAshi\\
Fundamental operation & mUlaBUta parxtikirxye\\
Generalisation & sAvaRtirxVkaraNa\\
Geometry & jAyxmiti, reVKAgaNita\\
Geometric progression & guNoVtatxra sherxVDhi\\
Hexahedron (Cube) & karxmaSaSaThx Palaka\\
Hindu Arabic numeral & hiMdU arebikf saMKAyxsUcakagaLu\\
Hypotenuse & vikaNaR\\
Icosahedron & karxmaviMshati Palaka, viMshati muKa Gana\\
Imaginary number  & kAlapxnika saMKeyx; UhA saMKeyx\\
Index & GAtasUci\\
Indeterminate & anidhARrita\\
Indeterminate equation of & modalaneGAtada anidiRSaTx\\[-0.1cm]
the first degree & samiVkaraNa\\
Indeterminate equation of & eraDaneya GAtada\\[-0.1cm] 
the second degree &  anidiRSaTx samiVkaraNa\\
Inductive & nigamana\\
Inference & tiVmARna\\
Integer & pUNARMka\\
Interpolation & aMtaHkeSxVpa\\
Inverse & parxtiloVma\\
Irrational & aBAgalabadhx\\
Isosceles & samadivxbAhu\\
Isosceles Triangle & samadivxbAhu tirxBuja\\
Logic & takaRshAsatxrX\\
Mathematics & gaNitashAsatxrX\\
Mathematician & gaNitajacnx\\
Magic square & mAyAcwka\\
Mensuration & keSxVtarxgaNita\\
Mixed fraction & misharxBinanxrAshi\\
Multiptier & guNaka\\
Multiply & guNisu\\
Natural numbers & sAvxBAvika saMKeyxgaLu\\
Negative number & QuNasaMKeyx\\
Negative sign & QuNacihenx\\
Number Theory & saMKAyxsidAdhxMta\\
Numeral & saMKAyxsUcaka\\
Obeservatory & viVkaSxNAlaya\\
Octahedron & karxma aSaTxPalaka; aSaTxmuKa Gana\\
Palindrome & mAlA saMKeyxgaLu\\
Paradox & viroVdhABAsa\\
Parallel lines & samAnAMtara reVKegaLu\\
Pascals triangle & pAsakxlana tirxBuja\\
Perfect number & paripUNaR saMKeyx\\
Permutation & karxmayoVjane\\
Place value & sAthxnabele\\
Positive number & dhanAMsha\\
Positive integer & dhana pUNARMsha\\
Probabality & saMBavaniVyate\\
Progression & sherxVDhi\\
Project & yoVjane\\
Prime number & aviBAjayx saMKeyx\\
Pythagorian tripletes & peYthAgoriyanf tirxvaLigaLu\\
Pythagoras theorem & peYthAgorasf parxmeVya\\
Puzzle & samaseyx\\
Quadratic equation & vagaRsamiVkaraNa\\
Quotent & BAgalabadhx\\
Queen of science & vijAcnxnada rANi\\
Ramanujam numbers & rAmAnujanf saMKeyxgaLu\\
Ratio & parxmANa\\
Rational Numbers & parimeVya saMKeyx; BAgalabadhx saMKeyx\\
Recurring & AvataR\\
Remainder & sheVSa\\
Resolve & viGaTisu\\
Result & PalitAMsha\\
Right triangle & laMbakoVna tirxBuja\\
Rule & niyama\\
Rule of three & terxYrAshi\\
Semi Circle & adhaRvaqtatx\\
Sign & cihenx\\
Similarity & sAmayx\\
Simple equation & saraLa samiVkaraNa\\
Simplification & saraLiVkaraNa\\
Simplify & saMkeSxVpisu\\
Solution & parihAra\\
Sphere & goVLa\\
Square & vagaR, cwka\\
Square root & vagaRmUla\\
Substitute & AdeVshisu\\
Subtraction & vayxvakalana\\
Tetrahedron & karxma catuSaPxlaka ; catuRmuKa Gana\\
Theorem & parxmeVya\\
Theory & sidAdhxMta\\
Theory of Indicies & GAtAMka tatavx\\
Theory of Numbers & saMKAyxshAsatxrX ; saMKAyxsidAdhxMta\\
Theory of Probability & saMBavaniVyate sidAdhxMta,\\[-0.1cm] 
 &saMBavayxtA sidAdhxMta\\
Trignometry & tirxkoVnamiti\\
Transpose & sathxLAMtarisu\\
Unique & EkeYka ; ananayx\\
Vedic mathematics & veVdagaNita\\
Verify & tALenoVDu\\
Whole number & pUNaRsaMKeyx\\
Zero & sonenx ; shUnayx
\end{longtable}

\newpage
\section*{AdhAra garxMthagaLu}
\addcontentsline{toc}{section}{{\protect\bf AdhAra garxMthagaLu}}

`gaNita veYvidhayx' pusatxkakAkxgi I muMde namUdisiruva garxMthagaLa sahAyavanunx paDedidedxVne. pusatxkada leVKakarigU parxkAshakarigU nananx kaqtajacnxtegaLu.

\begin{longtable}{l>{\raggedright}p{4cm}>{\raggedright}p{4cm}}
{\rm 1)} &  gaNitashAsatxrX cariterx & DA| si.~enf. shirxVnivAsayayxMgArf \tabularnewline
{\rm 2)} & shirxVnivAsa rAmAnujanf & DA| esf. bAlacaMdarxrAvf\tabularnewline
{\rm 3)} & shirxVnivAsa rAmAnujanf & porx| ji.~Ti. nArAyaNarAvf\tabularnewline
{\rm 4)} & vijAcnxna kaNARTaka & meYsUru vishavxvidAyxnilaya\tabularnewline
{\rm 5)} & catuvaRNaR samaseyx & porx. ji.~Ti. nArAyaNarAvf\tabularnewline
{\rm 6)} & gaNitashAsatxrXda parxvataRkaru & DA| esf. bAlacaMdarxrAvf\tabularnewline[-0.1cm]
& matutx sAvxrasayxgaLu & emf. sheYlaja, vi. vanaja\tabularnewline
{\rm 7)} & mAyAcwkagaLa sAvxrasayx & DA| esf. bAlacaMdarxrAvf\tabularnewline
{\rm 8)} & aMki saMKeyxgaLa sAmArxjayxdalilx & bi.~ke. vishavxnAtharAvf\tabularnewline
{\rm 9)} & BAratada parxsidadhx gaNitajacnxru  & saMpAdakaru \tabularnewline[-0.1cm]
& ({\rm 19-20} neV shatamAnada kiVtiRsheVSaru) & DA| esf. bAlacaMdarxrAvf\tabularnewline
{\rm 10)} & mAyA cwkagaLa mAyA parxpaMca & bi.~ke. vishavxnAthfrAvf\tabularnewline
{\rm 11)} & {\rm History of Mathematics} & {\rm D. S. Smith (part 1 and 11)}\tabularnewline[-0.1cm]
{\rm 12)} & kananxDa vishavxkoVsha &\tabularnewline
{\rm 13)} & KagoVLashAsatxrX & DA|~si.~enf. shirxVnivAsayayxMgArf\tabularnewline
{\rm 14)} & aMtajARlagaLu-gUgalf vikipiVDiyA &\tabularnewline
\end{longtable}

\newpage
\section*{ideV leVKakara parxkaTita kaqtigaLu}
\addcontentsline{toc}{section}{{\protect\bf ideV leVKakara parxkaTita kaqtigaLu}}

\begin{enumerate}
\item vijAcnxna keYpiDi (janapirxya vijAcnxna)

\item vijAcnxna oMdu pakiSxnoVTa

\item vijAcnxna veYvidhayx

\item vijAcnxna keYpiDi (pwrxDhashAlA vidAyxthiRgaLige)

\item AroVgayx-parisara-paradhamaR

\item gaNita mAdhuyaR

\item meduLige kasaratutx (gaNitada samaseyxgaLu)

\item gaNita agaNita

\item saMKAyxvinAyxsa matutx aviBAjayx saMKeyxgaLu

\item nAveSuTx budidhxvaMtaru (gaNitakekx saMbaMdhisidaMte)

\item makakxLa keY caLaka (reVKAgaNitakekx saMbaMdhisidaMte)

\item jagatitxna parxKAyxta gaNitajacnxru

\item saMKAyx parxpaMca

\item gaNitada nUtana mAdari parxshenxgaLu
\end{enumerate}

