\chapter{AdhAra garxMthagaLu}


\begin{longtable}{l>{\raggedright}p{4cm}>{\raggedright}p{4cm}}
{\rm 1)} &  gaNita shAsatxrX cariterx & DA| si.enf. shirxVnivAsayayxMgArf \tabularnewline
{\rm 2)} & shirxVnivAsa rAmAnujanf & DA|esf. bAlacaMdarxrAvf\tabularnewline
{\rm 3)} & shirxVnivAsa rAmAnujanf & porx| ji.Ti. nArAyaNarAvf\tabularnewline
{\rm 4)} & vijAcnxna kaNARTaka & meYsUru vishavxvidAyxnilaya\tabularnewline
{\rm 5)} & catuvaRNaR samaseyx & ji.Ti. nArAyaNarAvf\tabularnewline
{\rm 6)} & gaNitashAtxrXda parxvataRkaru & DA| esf. bAla caMdarxrAvf\tabularnewline[-0.1cm]
& matutx sAvxrasayxgaLu & emf. sheYlaja, vi. vanaja\tabularnewline
{\rm 7)} & mAmAcwkagaLa sAvxrasayx & DA| esf. bAlacaMdarxrAvf\tabularnewline
{\rm 8)} & aMki saMKeyxgaLa sAmArxjayxdalilx & bi.ke. vishavxnAtharAvf\tabularnewline
{\rm 9)} & BAratada parxsidadhx gaNitajacnxru  & saMpAdakaru \tabularnewline[-0.1cm]
& ({\rm 19-20} neV shatamAnada kiVtiRsheVSaru) & DA| esf. bAlacaMdarxrAvf\tabularnewline
{\rm 10)} & {\rm Vedic Mathematics} & {\rm Jagadguru Swami Bharatikrishna}\tabularnewline[0.1cm] 
&& {\rm  Tirthaji Maharaja}\tabularnewline
{\rm 11)} & mAyA cwkagaLa mAyA parxpaMca & bi.ke. vishavxnAthfrAvf\tabularnewline
{\rm 12)} & {\rm Histiry of Mathematics} & {\rm D.s. Smith (part 1 and 11)}\tabularnewline
{\rm 13)} & kananxDa vishavxkoVsha &\tabularnewline
{\rm 14)} & KagoVLashAsatxrX & DA| si.enf. shirxVnivAsayayxMgArf\tabularnewline
{\rm 15)} & veVdagaNita & DA| si. enf. shirxVnivAsa ayayxMgArf\tabularnewline
{\rm 16)} & {\rm The History of Ancient} & DA| si.enf. shirxnivAsayayxMgArf\tabularnewline[-0.1cm]
& {\rm Mathematics}& \tabularnewline
{\rm 17)} & aMtajARlagaLu-gUgalf vikipiVDiyA & \tabularnewline
\end{longtable}

