\makeatletter
\def\@makeschapterhead#1{%
  \vspace*{50\p@}%
  {\parindent \z@ \raggedright
    \normalfont
    \interlinepenalty\@M
    \LARGE \bfseries  #1\par\nobreak
    \vskip 40\p@
  }}
\makeatother

\chapter*{Acknowledgements}\label{ack}

\lhead[\small\thepage\quad Western Indology \& Its Quest For Power]{}
\rhead[]{Acknowledgements\quad\small\thepage}


The organisers of the conference are indebted to the help rendered by various 
institutions and individuals. It is a pleasure to thank them heartily for the same.

We are grateful to IIT-Madras and IGNCA, New Delhi for providing us the venue for the two conferences.
In particular, we are thankful to Prof. Jalihal, Sri Udaykumar and his team for their help in making 
full arrangements for the first conference at Chennai, and to Sri Sachchidanand Joshi, Sri Sudhir Lall, 
Sri Seetharamiah and their team who put in enormous efforts to make the conference at Delhi a success. 
Sri Jithu Aravamudhan and Smt. Lakshmi Sarma of IFI group of volunteers also deserve our hearty thanks 
for their active participation. Ms. Ruchi Sood and her troop of volunteers as also Sri Roushan Rajput 
and Sri Divyanshu Bawa made no small efforts in making the New Delhi conference a success.

Our gratitude is due to Sri Ramnik Khurana, Sri Sanjeev Chhibber, and 
Sri Chetan Handa who have kindly offered to bear the expenses of bringing out these volumes in print.
We wish to thank Sri Sunil Sheoran who has been a long time supporter of our work. His help in coordinating the
printing of these volumes is deeply appreciated.

We are grateful to all the paper-presenters and the keen audience for maintaining a high academic standard 
and decorum at the conference. There of course are many more who have helped us and guided us behind the 
screen and deserve our thanks.
\bigskip

~\hfill {\bf IFI Team}
