\chapter*{About Infinity Foundation India}\label{preface4}

\vspace{-.7cm}

\begin{center}
\dev{॥ इच्छन्ति देवाः सुन्वन्तं} \\
 \dev{न स्वप्नाय स्पृहयन्ति॥}
\end{center}

\begin{center}
“The deva-s love the performer of yajña,\\
not the one who slumbers” — Ṛgveda 8.2.18
\end{center}


Infinity Foundation India (IFI) is a public charitable and educational trust based out of Chennai (India) engaged in education and research in the area of civilization studies. Our vision is to bring about advancement of national integrity, education and the building of an Indian civilizational narrative. We celebrate \textbf{being different} and look forward to a world where myriad cultures thrive in an atmosphere of \textbf{mutual respect}. 

Infinity Foundation India is at the forefront of civilization studies from an Indian point of view, and strives to ensure that Indian knowledge systems are given their due as legitimate cultural assets on the world stage. Our range of activities are broad, and include education and research, media engagement, policy recommendations for various public and private bodies, and social activism for the betterment of society. There are four tracks of activities:

\begin{itemize}
\itemsep=0pt
\item Original R\&D/book publishing on various topics on Indian civilization and culture

\item Swadeshi Indology Conferences: These conferences are intended to take up the subject matters and produce more scholarship and publish more books. The idea is to build a Home Team of intellectuals of the highest standard.

\item E-learning: Through this track, IFI develops courses based on topics where it has research leadership.

\item Outreach. This includes popular level mass distribution, using media and social media.

\end{itemize}

\vspace{-.3cm}

One of the goals of Infinity Foundation India is to develop, fund, and groom scholars who can methodically respond to the Western worldview of Indology. We have conducted high impact conferences in the past with quality output for publications. Our areas of research include history, science, social and political thought, mind and body healing, and spirituality. Our Intellectual Kshatriya portal provides a platform where scholars and public intellectuals write on contemporary issues and modern narratives from an Indian civilizational standpoint. We also conduct internship and educational programs for interested students and scholars of Indian civilization.

\textbf{Infinity Foundation India Conferences}


Infinity Foundation India conferences have been envisioned to counter the 250-year-old narrative of Western Indology, the genre of Orientalism that focuses on India. As is well known, Western Indology helped formulate and legitimize the British colonial policies and played a very crucial role in the successful colonization and oppression of the Indian peoples. Our conferences and publications from them intend to analyze and respond to the well-orchestrated, systematic attack on:
\begin{itemize}
\itemsep=0pt
\item the foundational elements (the Vedas, Upanishads, Itihasas, and Puranas),

\item the living principles (the purusharthas and the varnashrama system),

\item the cultural manifestations (shastras, kavyas, and kalas) and

\item the sacredness of the still living articulations (art, temples, Kumbha-mela, festivals, family and social-structures)
\end{itemize}

From 2016 till date, IFI has conducted 6 Swadeshi Indology conferences on various issues pertaining to the advancement of national integrity, education and the building of an Indian civilizational narrative.

In future, Infinity Foundation India plans to conduct numerous high impact conferences on the History of Indian Science and Technology (HIST) in order to help promote the traditional Indian knowledge systems. This would also make the world appreciate how, in addition to spirituality and profound philosophy, scientific thought also has been inextricably linked to the development of ancient Indian civilisation. Such conferences would also demonstrate how the conceptual approach in Indian science has been different from that which has characterised the western development. Infinity Foundation India would also publish conference journals to promote further research and scholarship on this subject. We intend to develop E-learning courses on this subject for those interested. Besides, we would also develop high quality educational video content in the form of documentaries and short videos for non-specialists and specialists alike.

\label{endpreface4}
