%THE PREAMBLE
\documentclass[10pt]{article}
\usepackage{avg2}
\usepackage[pdfencoding=auto, psdextra]{hyperref}
\parskip 0.7\baselineskip
\raggedright

\begin{document}

\title{\bf On the Newtonian spacetime and a                
4-vector formulation of the Newton law of motion}           
 
   

\author{A.V.Gopala Rao,                                  
Department of  Studies in Physics,\\ University of Mysore, 
Manasagangotri,\\  Mysuru 570\,006, INDIA}
\maketitle
\begin{abstract} 
We discuss a tensor-geometric formulation of the 
Newton law of motion. 
\end{abstract} 

\begin{quote}
\textsl{\small Professor G Ramachandran was my senior 
colleague at the Department of Physics, 
University of Mysore. With pleasure and humility, 
I dedicate this article to the memory of the learned 
Professor.}
\end{quote}

\textsl{Notation}: Latin suffixes  $i,j,k,\dt$, are used 
for 
the spacetime range 0,1,2,3 and Greek suffixes  $\alpha, 
\beta, \gamma, \dt$,  for the spatial  range 1,2,3. Special 
upper case bold-symbols such as $\tns{T}, \tns{N}, 
\tns{U},\tns{A},\dt$,  denote 4-vectors in the Newtonian 
spacetime \mbb{N}.  Spatial-3-vectors are denoted by an 
over-head arrow as in $\veca{r}$,  $\veca{v}$,  $\veca{a}$, 
$\veca{f}$ etc. Bold-upper-case characters such as 
$\bmsf{T,N,U,A,F},\dt$, denote 4-dimensional matrices and 
4-dimensional column-matrices. Bold-lower-case characters 
such as $\bmsf{u,a,p,f},\dt$, denote 3-dimensional matrices 
and 3-dimensional column matrices.  An over-head ``tilde'' 
as in $\wt{\bmsf{A}}$ or  $\wt{\bmsf{a}}$ denote the 
transposes 
of the matrices  $\bmsf{A}$ and $\bmsf{a}$. Superscripts 
and 
subscripts, as in  $T^i$ and  $N_i$, are contravariant and 
covariant indices. A semicolon as in $N_{i;j}$ indicates  
covariant differentiation with respect to the 
\textsl{affinity} in the Newtonian spacetime \mbb{N}. 

\section{Introduction} 
The Minkowskian nature of the geometry of the spacetime of 
the special theory of relativity, and the general 
Riemannian 
nature of the geometry of the spacetime of the general 
theory of relativity, are well known. This makes a 
mathematically inclined student of physics  wonder if 
absolute space and absolute time which occur as independent 
geometric entities in Newtonian Mechanics, may similarly be 
combined together into a 4-dimensional geometric continuum, 
and if so, what could be the geometric nature of such a  
Newtonian spacetime. As an  answer to that query, we present 
in this article elements of the geometry of the Newtonian 
spacetime as discussed in Anderson's book [1] and, as an 
application of the method, discuss, very briefly, the 
Newtonian mechanics of a particle in accelerated frames of 
reference.

\section{The Newtonian spacetime} 
The geometry of the \textsl{Newtonian spacetime}  
$\mathbb{N}$ is described in detail in Anderson's book [1, 
pp.106-110]. Here, we quote  the essential elements of that 
discussion.

The \textsl{Newtonian spacetime} is a real 4-dimensional 
manifold  $\mathbb{N}$ endowed with a \textbf{symmetric flat 
affinity\footnote{The affinity is flat in the sense that the 
associated Riemann tensor is zero on $\mbb{N}$.}} 
$\bmsf{\Gamma}$, a covariant 4-vector field $\tns{N}$,  and 
a contravariant 4-vector field  $\tns{T}$. These objects  
$\bmsf{\Gamma}$, $\tns{N}$ and $\tns{T}$, are assumed to be 
defined   at all points of $\mathbb{N}$. As in relativity 
theory, points of $\mathbb{N}$ may be  called 
\textsl{events}. 

Furthermore, it is  assumed that the vector fields $\tns{N}$ 
and $\tns{T}$ satisfy the conditions 
\begin{align}\label{1} 
T^i N_i =1,\quad  N_{i;j}=0, \quad  T^i_{\;;j}=0, 
\quad  \partial_i  N_j - \partial_j N_i=0,
\end{align}
at every event of \mbb{N}. Above,  the partial  and 
covariant derivatives are defined in the usual way by
\begin{align*}
& \partial_i  N_j\equiv \partial N_j/\partial x^i,\quad 
T^i;j\equiv\partial_j  T^i+\Gamma^i_{kj} T^k,\quad 
N_{i;j}\equiv\partial_j  N_i-\Gamma^k_{ij} N_k,
  \end{align*}
where $\Gamma^k_{ij}$ are the coefficients of the symmetric 
affinity $\bmsf{\Gamma}$ in the coordinate system 
$\{x^i\}$. 
From Eqn.\eqref{1}, it is clear that $\tns{N}$ and 
$\tns{T}$ are covariantly constant vector fields. Further, 
one can introduce [Anderson,1] a \textsl{global canonical 
coordinate System} 
\begin{align}\label{2}
 \{x^i\}\equiv \{x^0\equiv t, x^1\equiv 
x,x^2\equiv y,x^3\equiv z\} 
\end{align}
in  \mbb{N}  in which 
\begin{align}\tag{2b}
\Gamma^i_{jk}=0, 
\quad 
T^i=(1,0,0,0),
\quad 
N_i=(1,0,0,0),
\end{align}
\textbf{at all events of} \mbb{N}. Observe that   the 
covariant components $N_i$ of $\tns{N}$ given in Eqn.(2b) 
may be expressed as  
\begin{align*}\tag{2c}
N_i=t_{;i} \equiv (\partial t/\partial 
x^i)=(1,\;0,\;0,\;0).
\end{align*}
This is the necessary and sufficient condition for 
$\tns{N}$ to be  a \textsl{hypersurface orthogonal} 
4-vector 
field:  Clearly, in the canonical coordinate system 
$\{x^i\}$  in Eqn.(2), $N_i$ is normal to the 
3-dimensional 
hypersurfaces $t=\textsl{constant}$. The 3-dimensional 
hypersurfaces  $t=\textsl{constant}$  are called the 
\textbf{hypersurfaces of absolute simultaneity} in \mbb{N}. 
All events on such a hypersurface of absolute simultaneity 
occur at the same instant  $t$. 

\subsection{Inertial frame}
The \textsl{global, canonical coordinate system} 
$\{x^i\}\equiv\{t, x,y,z\} $ mentioned above in Eqn.(2) has 
space-coordinates $\{x,y,z\}$ which are Cartesian. Further, 
in this canonical coordinate system in Eqn.\eqref{2}, in 
which the properties (2b) and (2c) are true, all the 
components of affine connection $\Gamma^i_{jk}$ are    zero 
(showing that this  symmetric affinity \bmsf{\Gamma} is 
indeed flat). This global, canonical coordinate system in 
\mbb{N} is called an \textbf{inertial coordinate system} 
or, 
an \textbf{inertial frame}. 

It is useful to note that in an inertial frame, the time 
and 
space intervals between two events $P:(t_P,x_P,y_P,z_P)$ 
and $Q:(t_Q,x_Q,y_Q,z_Q)$  are given by
\begin{align}\label{3}
(\Delta  t_{PQ})^2=(t_P-t_Q)^2,
\end{align}
and 
\begin{align}\label{4}
(\Delta  \ell_{PQ})^2=(x_P-x_Q)^2+(y_P-y_Q)^2+(z_P-z_Q)^2.
\end{align}

\subsection{Non-inertial frame} 
In addition to an inertial frame $\{x^i\}\equiv\{t,x,y,z\}$, 
coordinate frames  $\{\bar{x}^i\}$ obtained from an 
inertial frame by a transformation of the type 
\begin{align}\label{5}
\bar{x}^i=\bar{x}^i(t, x,y,z),
\end{align}
where $\bar{x}^i (t, x,y,z)$ are arbitrary differentiable 
functions of the $(t, x,y,z)$,  are also of interest in 
physics. Of particular importance among these are a 
sub-class of transformations of the type\footnote{Since we 
confine ourseleves to considering coordinate 
transformations of the type (5a) which preserve the 
absolute time $t$, we write, in what follows   
$\bar{x}^0=x^0=t$ to simplify the notation.} 
\begin{align*}\tag{5a}
& \bar{x}^0= t,\\
& \bar{x}^\alpha=\bar{x}^\alpha(t, x,y,z),\quad 
\alpha=1,2,3,
\end{align*}
and this class includes most of the  frames which are of 
interest in physics. We note that the \textbf{Galilei 
transformation} 
\begin{align}\label{6}
 \bar{x}^0=t,\;\veca{\bar{r}} 
=\veca{r}-\veca{v_0}t ,
\end{align}
where $\veca{{r}}\equiv  
(x,y,z)$, $\veca{\bar{r}}\equiv  
(\bar{x}^1,\bar{x}^2,\bar{x}^3)$ and $\veca{v_0}\equiv  
(v^1,v^2,v^3)$ is a constant 3-vector, belongs to this 
class defined in Eqn.(5a). A Galilei transformation, 
however,  transforms an  inertial frame $\{x^i\}$ into 
another inertial frame $\{\bar{x}^i\}$. Therefore, if our 
interest lies in transforming an inertial frame into a 
non-inertial frame, we must consider non-Galilei 
transformations. 



\section{4-trajectory, 4-velocity and  4-acceleration}
The \textbf{spacetime trajectory or 4-trajectory}  $C$ of 
a point-particle is a curve in  \mbb{N}. In a coordinate 
system   $\{x^i\}$,  with $x^0\equiv t$, $C$ is described 
by the parametric equation  $C: x^i=x^i(t)$. A general  
event on  $C$ at time $t$ would then have the 
coordinates  $(t,\;x^\alpha (t))$. 

The  \textbf{4-velocity} $\tns{U}$ of the particle is a 
\textbf{tangent vector} to  $C$. In dealing with the 
4-velocity, it is convenient to work with a 4-dimensional 
column-matrix $\bmsf{U}$ whose elements    are the 
contravariant components of $\tns{U}$, namely, $U^0=\dd 
x^0/\dd t=1$, $U^1=\dd x^1/\dd t$, $U^2= \dd 
x^2/\dd t$, $U^3=\dd x^3/\dd t$ in 
$\{x^i\}$. Therefore, we have
\begin{align}\label{7}
\tns{U}\mapsto{\bmsf{U}}
=(U^i)=(\dd x^i /\dd t)
=\begin{paray}{c} \dotup{x}{}^{0}\\ \dotup{x}{}^1\\ 
\dotup{x}{}^2\\\dotup{x}{}^3\end{paray}=  
\begin{paray}{c} 1\\ \bmsf{u}\end{paray},
\end{align}
where the derivatives $\dotup{x}{}^\alpha, \,\alpha= 1,2,3$,
are evaluated on $C$ and the 3-column matrix  
\begin{align}\label{8}
\bmsf{u}=\begin{paray}{c}u^1\\ u^2\\ 
u^3\end{paray} =\begin{paray}{c}\dot{x}{}^1\\ \dot{x}{}^2\\ 
\dot{x}{}^3\end{paray},
\end{align}
is called the 3-velocity of the particle. 

The \textbf{4-acceleration}   $\tns{A}$ of a particle is 
defined as the \textbf{intrinsic derivative} ${\mathrm{D}} 
\tns{U}/\dd t$ of the 4-velocity $\tns{U}$ along the 
4-trajectory $C$ of the particle. Therefore, the  
contravariant components of the 4-acceleration are 
\begin{align}\label{9}
A^i= {\mathrm{D}}U^i /dt=\dd  U^i /\dd t +
\Gamma^i _{jk}  U^j U^k ,
\end{align}
which involve the \textsl{affine-coefficients}  $\Gamma^i 
_{jk}$ in $\{x^i\}$. In particular, in an \textbf{inertial 
frame} $\{x^i\}\equiv\{t,x,y,z\}$, where $({\mathrm D}U^i
/\dd t)=    ( \dd U^i /\dd t)$ because  all the 
affine-coefficients vanish at all events, we have
\begin{align}\label{10}
{\bmsf{A}}=(A^i) =( \dd U^i /\dd t)
=\begin{paray}{c} 0\\ \ddot{x}
\\ \ddot{y}\\ \ddot{z}\end{paray}
 =\begin{paray}{c} 0\\ \bmsf{a}\end{paray}.
\end{align}
 The  3-dimensional column matrix $\bmsf{a}$ in 
Eqn.\eqref{10} 
above, is defined by  
\begin{align}\label{11}
\bmsf{a}=\begin{paray}{c}
a^1\\a^2\\a^3\end{paray}=\dotup{\bmsf{u}}=\begin{paray}{c}
\dot{u}^1\\\dot{u}^2\\\dot{u}^3\end{paray}
= \begin{paray}{c}\ddot{x}\\\ddot{y}\\\ddot{z}\end{paray}.
\end{align}

\section{4-Momentum, 4-Force and Newton's law of motion 
as a 4-vector law}
A point-particle is characterised 
by a \textbf{mass} $m$ which is a \textbf{4-scalar}.  
The \textbf{4-momentum} of a point-particle of mass  $m$ 
is defined  as $\tns{P}=m\tns{U}$. Therefore (see 
Eqn.\eqref{7} for ${\bmsf{U}}$),
\begin{align}\label{12}
{\bmsf{P}}=m{\bmsf{U}}=\begin{paray}{c}m \\ 
m\vec{u}\end{paray}.
\end{align}
Now, we may state Newton's (second) law of motion in \mbb{N} 
as the 4-vector equation 
\begin{align}\label{13}
{{\mathrm{D}}}\tns{P}/\dd t=\tns{F},
\end{align}
where $\tns{F}$ is the \textbf{4-force} acting on 
the particle $m$. Using the  corresponding column-matrices 
$\bmsf{P}$ and 
$\bmsf{F}$ in a given frame $\{x^i\}$, we may 
state Newton's law,  equivalently, as
\begin{align}\label{14}
{\mathrm{D}}{\bmsf{P}}/\dd t={\bmsf{F}}.
\end{align}

The 4-force $\bmsf{F}$ is assumed to be
a function of the coordinates and momentum of the particle. 
Explicitly, we may express  Newton's law Eqn.\eqref{14}  as 
\begin{align}\label{15}
\dd P^i /\dd t + \Gamma^i_{jk}P^j
U^k=F^i, \quad  U^k \equiv \dd x^k/\dd  t, \quad \text{and} 
\quad P^k \equiv m U^k.
\end{align}

\section{A  constraint on the 4-force}
We recall (see equations (2b) and \eqref{12})  that in  an 
\textbf{inertial frame} $\{x^i\}$, $N_i=(1,\vec{0})$ and 
$P^i=mU^i= (m,m\vec{u})$. Therefore,  the scalar  
$N_iP^i\;$ equals (the scalar) $m$ in  an inertial frame 
and hence in every (allowed) frame. Using this 
observation, we note that  \textbf{for a particle whose mass 
$m$ is a constant along its 4-trajectory $C$},
\begin{align*}
&0 = \dd m/\dd t = \dd (N_i P^i )/\dd t =
{\mathrm{D}}(N_i P^i )/\dd t  \notag \\  
&=({\mathrm{D}}N_i/\dd t)P^i + N_i ({\mathrm{D}}P^i/\dd t) 
= 
\underline{(N_{i;j}U^j)P^i 
} +N_i({\mathrm{D}}P^i/\dd t) \notag \\  
&= N_i({\mathrm{D}}P^i/\dd t) = N_i F^i,
\end{align*}
where the `underlined-term' has dropped  out because the 
covariant derivative $N_{i;j}$ is zero. (Here, it is useful 
to recall that $U^i$ is a tangent 4-vector to the 
trajectory 
$C$.), Thus, if we require that the  4-force  $F^i $  
satisfies the  invariant constraint
\begin{align}\label{16}
N_i F^i =0,
\end{align}
we are assured that in all (allowed) coordinate systems, the 
mass $m$ of a particle does not vary along its trajectory. 
In an \textbf{inertial frame}
where $N_i$ has the special form in Eqn.(2b), the
constraint \eqref{16} implies, 
\begin{align}\label{17}
F^i =(0, \veca{f}).
\end{align}
and hence, Newton's law assumes the 
form ${{\mathrm{D}}}\tns{P}/\dd 
t=m\,{{\mathrm{D}}}\tns{U}/\dd t $  which may also be 
written as
\begin{align}\label{18}
\boxed{m\tns{A}  =\tns{F}.}
\end{align}
In an \textbf{inertial frame} where $ A^0 =F^0=0$, the above
equation reduces to the familiar 3-vector equation
\begin{align}\label{19}
m\veca{a}  =\veca{f}.
\end{align}

In what follows, we assume that the 4-force 
satisfies the constraint \eqref{16}.

\section{An example: Motion relative to a 
non-inertial frame}

To illustrate  our  discussion, we  consider motion of a 
particle relative to an  \textsl{accelerated reference 
frame} $\{\bar{x}^i\}$ obtained by  a
coordinate 
transformation of the type (5a): 
\begin{align}\label{20}
\bar{t}=t, \quad \bar{x}^\alpha
=x^\alpha - q^\alpha(t), \quad \alpha=1,\;2,\;3.
\end{align} 
Here  $q^\alpha (t)$ are arbitrary differentiable functions 
of only $\bar{x}^0=x^0=t$. (Recall that  we write  
$\bar{x}^0=x^0=t$ to simplify the notation.) In a more 
familiar \textsl{3-vector notation}, this transformation 
\eqref{20} 
appears as:
\begin{align}\label{21}
\bar{t}=t, \quad  {\veca{\bar{r}}}=\veca{r} - 
\veca{q}(t),
\end{align}
where $\veca{q}(t)=(q^1(t),q^2(t),q^3(t) )$.

\subsection{The Jacobian matrices}
The forward and inverse Jacobian matrices of the 
coordinate transformation  Eqn.(20) are, respectively, 
% \begin{small}
\begin{align}\label{22}
\bmsf{J} &=(?J^i_j?)=\left(\frac{\partial 
\bar{x}^i}{\partial x^j}\right)
=\begin{paray}{cccc}1&0&0&0\\ 
-\dot{q}^{1}&1&0&0\\
-\dot{q}^{2}&0&1&0 \\
-\dot{q}^{3}&0&0&1 \end{paray}
=\begin{paray}{cc} 1 & \wt{\bmsf{0}}\\
-\bmsf{\dot{q}} & \bmsf{E}_3 \end{paray}, 
\end{align}
\begin{align}\label{23}
\bmsf{K}&=(?K^i_j?)=\left(\frac{\partial x^i}{\partial 
\bar x^j}\right)=\begin{paray}{cccc}1&0&0&0\\ 
\dot{q}^{1}&1&0&0\\
\dot{q}^{2}&0&1&0 \\
\dot{q}^{3}&0&0&1 \end{paray}
= \begin{paray}{cc} 1 &\wt{\bmsf{0}}\\
\bmsf{\dot{q}} & {\bmsf{E}}_3 \end{paray}
=\bmsf{J}^{-1},
\end{align}
where, in the $4\times4$ matrices $(?J^i_j?)$ and 
$(?K^i_j?)$, the left  index $i$ is the row-index and the 
right index $j$ is the column-index.  We have also 
introduced $(1+3)$-block-matrices in the above two equations 
which are convenient in the calculations to follow. In these 
block-matrices,  $\wt{\bmsf{0}}$ is the row-matrix 
$(0\;0\;0)$, $\bmsf{E}_3$  is the $3\times 3$ unit matrix 
and  $\bmsf{\dot{q}}$ is the 3-dimensional column matrix
\begin{align*}\tag{23a}
\bmsf{\dot{q}}=\begin{paray}{c}\dot{q}^{1}\\ \dot{q}^2 \\
\dot{q}^3\end{paray}.  
\end{align*}
Also, while  calculating the partial derivatives  $?J^i_j?$ 
in the Jacobian ${\bmsf{J}}$, one must remember that $x^0, 
x^1, x^2, x^3$ are \textbf{independent} variables, so that, 
for instance, 
\begin{align*}
 ? J^1_0?\equiv\partial \bar{x}^1/ \partial x^0
 =\partial(x^1-q^1)/ \partial t
 = -\partial q^1/\partial t=-\dot{q}^1,
\end{align*}
where $\partial x^1/\partial t=0$ because $x^1$ and $t$
are independent variables. Similarly, one may note that  
$\bar{x}^0,\; \bar{x}^1,\;  \bar{x}^2,\;  \bar{x}^3$ are 
\textbf{independent} variables while calculating 
$?K^i_j?$.

\subsection{N-vector}
We may express the covector transformation law
\begin{align*}\tag{24a}
\bar{N}_i=\frac{\partial x^j}{\partial \bar{x}^i}N_j=
(J^{-1})^j_{\;i} N_j
=\sum_{j=1}^3 (\wt{J}^{-1})^i_{\;j} N_j,
\end{align*}
conveniently as the $4\times4$ matrix equation 
\begin{align*}\tag{24b}
\bmsf{\bar{N}}= \wt{\bmsf{J}}^{-1} \bmsf{N}. 
\end{align*} 
We note that the matrix ${\bmsf{J}^{-1}}$ is given 
in Eqn.\eqref{23} and  the column matrix $\bmsf{N}$ in the 
inertial frame is that given in Eqn.(2b).  Plugging these  
into Eqn.(24b), we obtain
\begin{align}\label{24}
\bmsf{\bar{N}}=\wt{\bmsf{J}}^{-1} \bmsf{N}
= \begin{paray}{rc} 1 &  \wt{\bmsf{\dot{q}}} \\
\bmsf{0} & \bmsf{E}_3 \end{paray} 
\begin{paray}{rr} 
1\\\bmsf{0}\end{paray}
=\begin{paray}{c}  1\\{\bmsf{0}} \end{paray}
=\bmsf{N},
\end{align}
which shows that the normal vector to the 3-hypersurfaces  
of absolute simultaneity remains unaltered under the 
transformation \eqref{20}.

\subsection{4-velocity}
The contravector transformation law  $\bar{U}^i=(\partial 
\bar{x}^i/\partial x^j) U^j$ may similarly be written as 
the 
matrix equation  $ \bmsf{\bar{U}}=\bmsf{J} \bmsf{U}$ and  
we obtain (see Eqn.\eqref{22})
\begin{align}\label{25}
\bar{\bmsf{U}}=\bmsf{J}{\bmsf{U}}
=\begin{paray}{cc} 1 & \wt{\bmsf{0}}\\
-\bmsf{\dot{q}} & \bmsf{E}_3 \end{paray}
\begin{paray}{c} 1 \\ \bmsf{u} \end{paray}=\begin{paray}{c} 
1 \\ \bmsf{u}-\bmsf{\dot{q}} \end{paray}.
\end{align}

\subsection{4-momentum} 
Multiplying Eqn.\eqref{25} by  the (scalar) mass $m$ 
of the particle, we get the \textbf{4-momentum} 
transformation
\begin{align}\label{26}
\bar{\bmsf{P}}=\bmsf{J}{\bmsf{P}}
=\begin{paray}{cc} 1 & \wt{\bmsf{0}}\\
-\bmsf{\dot{q}} & \bmsf{E}_3 \end{paray}
\begin{paray}{c} m \\ m\bmsf{u} 
\end{paray}=\begin{paray}{c} 
m\\ m(\bmsf{u}-\bmsf{\dot{q}}) \end{paray}.
\end{align}

\subsection{Affinity}
In an inertial frame $\{x^i\}$ of \mbb{N}, the symmetric 
flat affinity $\Gamma^i_{jk}$ vanishes identically. In 
any other frame $\{\bar{x}^i\}$ obtained from an
inertial frame $\{x^i\}$ by a coordinate 
transformation $\bar{x}^i=\bar{x}^i(x^k)$,  the 
affinity $\bar{\Gamma}^i_{jk}$ is given by 
its  transformation law
\begin{align}\label{27}
\bar{\Gamma}^i_{jk}=\frac{\partial \bar{x}^i}{\partial
x^a} \frac{\partial}{\partial \bar{x}^j}\frac{\partial x^a}
{\partial \bar{x}^k}.
\end{align}
This transformation law may be  be expressed as a matrix 
equation as follows: For a \textbf{fixed} value of  $j$, we 
define two matrices  $\bar{\,\,\bsymb{\Gamma}_j} $ and  
${\bmsf{M}}_j$ in which the $i-$th row, $k-$th column 
elements are
\begin{align}\label{28}
(\bar{\bsymb{\Gamma}}_j)^i_{\;\,k}=\bar{\Gamma}^i_{jk} , 
\quad
({\bmsf{M}}_j )^i_{\;\,k}= \frac{\partial}{\partial
\bar{x}^j}\frac{\partial x^i} {\partial \bar{x}^k}
=\frac{\partial}{\partial \bar{x}^j} 
({\bmsf{J}^{-1}})^i_{\;\,k}.
\end{align}
Then observe that Eqn.\eqref{27} may be read as the  
four matrix equations  
\begin{align*}\tag{28a}
\bar{\bsymb{\Gamma}}_j= {\bmsf{J}}{\bmsf{M}}_j,\quad 
j=0,\;1,\;2,\;3.  
\end{align*}
Next, observe that in the coordinate transformation 
\eqref{20}, the elements of the matrix  $\bmsf{J}^{-1}$ are 
either constants or  functions of  $\bar{x}^0=t$ only (see 
Eqn.\eqref{23}). Hence the three matrices  ${\bmsf{M}}_1$, 
${\bmsf{M}}_2$ and  ${\bmsf{M}}_3$ in Eqn.\eqref{28}, 
because they involve differentiations with respect to $x^1$, 
$x^1$, and $x^3$ respectively, vanish identically and only  
the matrix  ${\bmsf{M}}_0$ is non-zero. It is
is given by (see Eqn.\eqref{23})
\begin{align}\label{29}
{\bmsf{M}}_0=\frac{\;\;\partial{\bmsf{J}}^{-1}}
{\partial\bar{x} ^0}
=\frac{\partial}{\partial t}\nnt
\begin{paray}{rc} 1 &\wt{\bmsf{0}}  \\ \bmsf{\dot{q}} 
 & \bmsf{E}_3 \end{paray}=\begin{paray}{rc} 0 
&\wt{\bmsf{0}} 
\\ \bmsf{\ddot{q}}  & \bmsf{0}_3 \end{paray},
\end{align}
where $\bmsf{0}_3$ is the $3\times3$ zero matrix. 
Finally, using the equations  (28a), \eqref{22}, and
\eqref{29}, we obtain
\begin{align}\label{30}
\bar{\bsymb{\Gamma}}_0= 
{\bmsf{J}}{\bmsf{M}}_0
=\begin{paray}{cc} 1 & \wt{\bmsf{0}}\\
-\bmsf{\dot{q}} & \bmsf{E}_3 \end{paray}
\begin{paray}{rc} 0 &\wt{\bmsf{0}} \\ 
\bmsf{\ddot{q}}  & \bmsf{0}_3 \end{paray}
=\begin{paray}{cc}0&\wt{\bmsf{0}}\\
\bmsf{\ddot{q}}& \bmsf{0}_3 \end{paray},
\end{align}
which  in expanded notation, is
\begin{align}\label{31}
\bar{\bsymb{\Gamma}}_0\equiv 
\begin{paray}{cccc}
\Gamma^0_{00}&\Gamma^0_{01}
&\Gamma^0_{02}&\Gamma^0_{03}\\
\Gamma^1_{00}&\Gamma^1_{01}
&\Gamma^1_{02}&\Gamma^1_{03}\\
\Gamma^2_{00}&\Gamma^2_{01}
&\Gamma^2_{02}&\Gamma^2_{03}\\
\Gamma^3_{00}&\Gamma^3_{01}
&\Gamma^3_{02}&\Gamma^3_{03}\end{paray}
= \begin{paray}{cccc} 
0&0&0&0\\
{\ddot{q}}^1&0&0&0\\
{\ddot{q}}^2&0&0&0\\
{\ddot{q}}^3&0&0&0
\end{paray}.
\end{align}
From this, we pick the following three non-zero components 
of the flat affinity in the accelerated frame  
$\{\bar{x}^i\}$:
\begin{align}\label{32}
\bar{\Gamma}^\alpha_{00}=\ddot{q}^\alpha,\quad 
\alpha=1,2,3.
\end{align}

\subsection{4-force} 
The contravariant components of the 4-force in the   
accelerated frame $\{\bar{x}^i\}$  are obtained through the 
 matrix equation
\begin{align}\label{33}
\bar{\bmsf{F}}\equiv \begin{paray}{c} \bar{F}^0 \\
\bar{\bmsf{f}} \end{paray}=
\bmsf{J}\bmsf{F}=\begin{paray}{cc} 1 & \tilde{\bmsf{0}}  \\
-{\bmsf{\dot{q}}}& \bmsf{E}_3 \end{paray} 
\begin{paray}{rc}0 \\{\bmsf{f}} \end{paray}= 
\begin{paray}{cc} 0 \\ {\bmsf{f}}
\end{paray}={\bmsf{F}},
\end{align}
where $\bmsf{F}$ is the 4-force in  the inertial frame and 
is given by Eqn.\eqref{17}. Thus,
\begin{align}\label{34}
\bar{F}^0={F}^0=0, \quad 
\bar{\bmsf{f}}=\bmsf{f},
\end{align}
showing that  the \textsl{4-force} is actually 
\textsl{invariant} under the map from an inertial frame to 
the  accelerated frame defined by the transformation 
\eqref{20}.

\subsection{Equation of motion}
The 4-vector Newton law in the 
accelerated frame  defined by the coordinate transformation 
\eqref{20}  consists of the following equations: 
\begin{align}\label{35}
&\frac{\dd m}{\dd t}=0,\\\label{36}
&\frac{\mathrm{D} {\bar{P}}^\alpha}{\dd t\;\:}= 
\frac{\dd \bar{P}^\alpha}{\dd t\;\:} + 
\bar{\Gamma}^\alpha_{00} \bar{P}^0 \bar{U}^0= 
{\bar{F}^\alpha},\quad 
\alpha=1,2,3. 
\end{align} 
Of these equations, the equation \eqref{35} says nothing 
new; it  is consistent with the postulate  that the mass 
$m$ 
of the  particle does not change along its trajectory. 
Using 
the 4-force in Eqn.\eqref{34}, and the affinity in 
Eqn.\eqref{32}, in Eqn.\eqref{36}, we finally get
\begin{align}\label{37}
m\frac{\dd \bar{u}^\alpha}{\dd t\;}+m{\ddot{q}}{}^\alpha= 
f^\alpha, \quad 
\alpha=1,2,3,
\end{align}
where $\bar{f}^{\alpha}={f}^{\alpha}$ are the components of 
the \textbf{applied 3-force}. 
With $\bar{a}^\alpha\equiv{\dd \bar{u}^\alpha/\dd t}$,  the 
above equation becomes
\begin{align}\label{38}
m\bar{a}^\alpha= f^\alpha+ 
f_{\rm{fict}}^\alpha,\quad \alpha=1,2,3.
\end{align}
where we have transferred the second term on the left side 
of Eqn.\eqref{37} to the right and defined
\begin{align}\label{39}
f_{\rm{fict}}^\alpha \equiv -m\ddot{q}^\alpha,\quad 
\alpha=1,2,3,
\end{align}
which are the components of a 3-vector called the 
\textbf{fictitious force} arising in the  accelerated frame 
$\{\bar{x}^i \}$. We may recall that we usually consider the 
case  when $\veca{q}(t)=\veca{\alpha} t^2/2$ where 
$\veca{\alpha}$ is a constant vector and in that  case, we 
get $\veca{f}_{\nnt\rm{fict}}=-m\veca{\alpha}$ which 
corresponds to  the familiar fictitious force arising in a 
\textbf{linearly} accelerated frame.

In books on mechanics, it is usual to consider next,  the 
problem of \textbf{a rigid, rotating, non-inertial  frame}  
$\{\bar{x}^i\}$. Such a frame may be obtained
from a inertial frame  $\{ x^i\}$ using the coordinate map:
\begin{align}\label{40}
\bar{x}^0=x^0, \quad
\bar{x}^\alpha=?{R}^{\alpha}_{\beta}?
?{x}^{\beta}_{}?, \quad \alpha=1,\;2,\;3,
\end{align}
where  $ ?{R}^{\alpha}_{\beta}?=
?{R}^{\alpha}_{\beta}? (t)$ 
are the elements of a (time-varying) rotation matrix 
${\bmsf{R}} (t)$ with elements $?{R}^{\alpha}_{\beta}?(t)$. 
Then, as is done in the example discussed above, one may 
calculate the affinity $\bar{\Gamma}^i_{jk}$ in the rotating 
frame $\{\bar{x}^i\}$, and obtain the fictitious forces 
which arise in that frame. However, we end our discussion 
here as our intention was only to introduce the geometrical 
aspects of the Newtonian spacetime and the  laws of motion. 
For more material on motion in a rotating frame, we refer 
the interested reader to standard treatises on 
classical mechanics.

\textbf{Acknowledgement} I wish to thank my 
colleague Professor K. S. Mallesh for several helpful 
observations and comments on the article.

\textbf{Reference}

[1] James L Anderson, 1967, Principles of relativity 
physics, Academic Press, New York and London, Sections 5-1 
and 5-2.

\end{document}

