\begin{thebibliography}{99}
\bibitem{key1} [abd] M. E. Abd El-Monsef, S. N. El-Deeb and R. A. Mahmoud, $\beta$-open sets and $\beta$ continuous mappings, Bull. Fac. Sci. Assut Univ., 12(1983), 77--90.
\bibitem{key2} [Abd1]A S Mashhour Abd M E El-Monsef and S N El-Deeb, On pre-continuous and weak pre continuous mappings, Proc. Math. Phys. Soc. Egypt, 53 (1982), 47--53.
\bibitem{key3}  [Ahmad] Ahmad Al-Omari and Mohd. Salmi and Md. Noorani, On Generalized b-Closed Sets, Bull. Malays.Math. Sci. Soc. (2) 32(1) (2009), 19--30.
\bibitem{key4}  [Anitha] M.Anitha and P.Thangavelu, On Pre-Generalized pre-regular closed sets, Acta Ciencia Indica, 31M(4)(2005), 1035--1040.
\bibitem{key5} [Ganster] I. Arockiarani, K. Balachandran and M. Ganster, Regular generalized locally closed sets and RGL-Continuous functions Indian J. Pure appl. Math., 28(5) (1997), 661--669.
\bibitem{key6} [Arockiarani] I. Arockiarani, Studies on generalizations of generalized closed sets and maps in topological spaces, Ph.D Thesis, Bharathiar University, Coimbatore, (1997).
\bibitem{key7} [Arya1] S.P. Arya and R. Gupta, On strongly continuous mappings, Kyungpook Math., J. 14(1974), 131--143.
\bibitem{key8} [Arya] S. P. Arya and T. M. Nour, Charactarization of S-normal spaces, Indian J. Pure Appl. Math  21(1990), 717--719.
\bibitem{key9}  [Arya2] S. P. Arya and T. M. Nour, Separation axioms for bitopological spaces, Indian J. Pure Appl. Math., 19(1998), 42--50.
\bibitem{key10} [Baker] Baker C W, Contra open and Contra Closed functions, Math. Sci., 1994, 17:413--415.
\bibitem{key11} [Basavaraj] Basavaraj M Ittanagi and Veeresha A Sajjanar ``On weakly semi closed sets in Topological spaces''  International Journal Of Mathematical Archive-8(9), 2017, 126--134 8(9), 126--134.
\bibitem{key12}  [Benchalli] S.S. Benchalli and R.S. Wali, On RW-closed sets in topological spaces, Malays. Math. sci. Soc. 30 (2007), 99--110.
\bibitem{key13} [Benchalli1] S.S. Benchalli, P.G. Patil and T.D. Rayanagaudar, $\omega\alpha$-Closed sets is Topological Spaces, The global j. Appl. Math. Sci. 2, 53--63.
\bibitem{key14} [Benchalli2] S.S. Benchalli, P.G. Patil and P.M. Nalwad. Genaralized  $\omega\alpha$-Closed sets is Topological Spaces, The journal of new results in science Number: 7, Year: 2014, Pages: 7--19.
\bibitem{key15} [Bhattacharya1] Sharmistha Bhattacharya, On Generalized Regular Closed Sets, Int. J. Contemp. Math.  
\bibitem{key16} [Bhattacharyya]	P. Bhattacharyya and B.K.Lahiri, Semi-generalized closed sets in topology, Indian J. Math., 29 	(1987), 376--382.
\bibitem{key17} [Biswas] N. Biswas, On characterization of semi-continuous functions, Atti Accad. Naz. Lincei Rend, Cl. Sci. Fis. Mat. Natur. 48(8)(1970), 399--402.
\bibitem{key18} [Bourbaki] N. Bourbaki, General topology Part I, Addison-Wesley, Reading, Mass (1966).
\bibitem{key19} [Cameron] D.E. Cameron, Properties of s-closed spaces Proc. Amer. Math. Soc., 72 (1978), 581--586.(regular semi open) Coimbatore (2009).
\bibitem{key20} [Crossley] S.G. Crossley and S.K. Hildebrand, Semi topological properties, Fund. Math., 74(1972), 233--254.
\bibitem{key21} [Devi] K.Nono, R.Devi, M.Devipriya, K. Muthukumaraswamy and H. Maki, On $\TSg\#\alpha$-closed sets and the Digital Plane, Bull. Fukuoka Univ. Ed. Part III, 53(2004), 15--24.
\bibitem{key22} [Dhanapakyam] C. Dhanapakyam, J. Subashini and K. Indirani, ON - $\beta wg$* set and continuity in topological spaces, IJMA-5(5), 2014, 222--233.
\bibitem{key23} [Dontchev1] J Dontchev, Contra continuous functions and Strongly s-closed spaces, Int. J. Math. Sci. 19(1996), 15--31.
\bibitem{key24} [Dontchev] J. Dontchev, On generalizing semi pre open sets, Mem. Fac. Sci. Kochi Univ. Scr. A. 
\bibitem{key25} [Dontchev2] J. Dontchev and T. Noiri, Quasi-normal spaces and g-closed sets, Acta Math. Hungar. 89(3)(2000), 211--219.
\bibitem{key26} [Fukutake2] T. Fukutake, On generalized closed sets in bitopological spaces, Bull. Fukuoka Univ. Ed. Part III, 35(1986), 19--28.
\bibitem{key27} [Fukutake] T. Fukutake, P. Sundaram and N, Nagaveni, On weakly generalized closed sets, weakly generalized continuous maps and Twg spaces in bitopological spaces, Bull. Fukuoka Univ. Ed. Part III, 48(1999), 30--40.
\bibitem{key28}  [Gnanambal] Y. Gnanambal, Studies on generalized pre-regular closed sets and generalization of locally closed sets, Ph.D, Thesis Bharathiar University, Coimbatore, (1998).
\bibitem{key29} [Gnanambal1] T. Sampoornam, Gnanambal Ilango and N. Selvanayaki, Generalized Pre-Semi Closed Sets. In Topological Spaces, International Journal of Mathematical Archive-4(6), 2013, 151--154.
\bibitem{key30} [Indirani] K. Indirani and G. Sindhu, On Regular Generalized Star B closed sets, IJMA. International Journal of general topology, 2(1)(2009), 21--30.
\bibitem{key31}  [Janaki1] A. Jayalakshmi  \& C. Janaki, On wgr$\alpha$-Closed Sets in Topological Spaces, International Journal of Mathematical Archive-3(6), 2012, 2386--2392 KBM  Journal of Mathematical Sciences and computer Applications 2010 1(1): 9--17.
\bibitem{key32} [Janaki2] Savithiri D and Janaki C,. On Regular Generalized closed sets in topological spaces.
\bibitem{key33} [Janaki3] C. Janaki and Renu Thomas, On R* closed sets in topological space s, IJMA-3(8), 2012, 3067--3074.
\bibitem{key34} [Jastad] O.N. Jastad, On some classes of nearly open sets, Pacific J. Math, 15(1965), 961--970.
\bibitem{key35} [Kelly] J.C. Kelly, Bitopological spaces, Proc. London Math. Soc., 13 (1963), 71--81.
\bibitem{key36} M. Vivek Prabu, Generalized $\zeta$*-closed sets in topological spaces, International journal of mathematical archive, 4(5)(2013), 274--279.
\bibitem{key37} [Levine1] N. Levine, Semi open sets and Semi Continuity in Topological spaces, Amer. Math. Monthly, 70(1963) 36--41.
\bibitem{key38} [Levine] N. Levine, Generalized closed sets in topology, Rend. Circ. Mat. Palermo,  19(2)(1970), 89--96.
\bibitem{key39} [Levine2] N. Levine, Strong continuity in topological spaces, Amer. Math. Monthly, 67(1960), 269.
\bibitem{key40} [Long] Long PE, Herington LL. Basic Properties of Regular Closed Functions, Rend. Cir. Mat. Palermo1978; 27:20--28. 
\bibitem{key41} [Maheshwari] S.N. Maheshwari and R. Prasad, on s normal spaces, Bull. Math. Soc. Sci. Math. R.S. Roumanie 22 (1978) 27--28.
\bibitem{key42} [Maki2] K. Balachandran, P. Sundaram and H. Maki, On generalized continuous maps in topological spaces, Mem. Fac. Kochi Univ. Math., 12(1991), 5--13. 
\bibitem{key43} [Maki3] H. Maki, R. Devi and K. Balachandran, Associated topologies of generalized $\alpha$-closed sets and $\alpha$-generalized closed sets, Mem. Fac. Sci. Kochi Univ. Ser. A. Math., 15 (1994), 51--63.
\bibitem{key44} [Maki4] R. Devi, H. Maki and V. Kokilavani, The Group Structure of \#g$\alpha$-Closed Sets in Topological Spaces.
\bibitem{key45} [Maki5] Balachandran K, Sundaram P, Maki H. On Generalized homeomorphisms in Topological Spaces, Fukuoka university, Ed., part III. 40(1991) 13--21.
\bibitem{key46} [Maki6] H. Maki, P. Sundaram and K. Balachandran, On generalized continuous maps and pasting lemma in bitopological spaces, Bull. Fukuoka Univ. Ed. Part-III, 40(1991), 23--31.
\bibitem{key47} [Maki10] R. Devi, K. Balachandran and H. Maki, Generalized $\alpha$ closed maps and $\alpha$ generalized closed maps, Indian J. Pure Appl. Math., 29(1): 37--49, January 1998.
\bibitem{key48} [Maki11] R. Devi, K. Balachandran and H. Maki, On generalized $\alpha$ continuous maps, Far East J. Math. Sci. Spacial volume, Part 1 (1997), 1--15.
\bibitem{key49} [Maki12] R. Devi, K. Balachandran and H. Maki, Semi generalized homeomorphisms and generalized semi homeomorphisms in topological spaces, Indian J. Pure Appl. Math, 26(1995), 271--284.
\bibitem{key50} [Sundaram] P. Sundaram, Studies on generalizations of continuous maps in topological spaces, Ph.D, Thesis, Bharathiar University, Coimbatore, (1991).
\bibitem{key51} [Balachandran] K. Balachandran, Y. Gnanambal and P. Sundaram, On generalized locally semi-closed sets and GLSC-continuous functions, Far East J. Math., Sci Special Volume (1997), part II 189--200.
\bibitem{key52} [Malghan] S. R. Malghan, Generalized closed maps, J. Karnatak Univ. Sci., 27(1982), 82--88.
\bibitem{key53} [Mariappa] K. Mariappa, S. Sekar, On Regular Generalized b Closed Set, Int. Journal of Math. Analysis, 	Vol. 7, 2013, no. 13, 613--624. Math. Soc., 41 (1937), 374--481.
\bibitem{key54} [Nagaveni] C. Mukundhan and N. Nagaveni, On semi weakly g* continuous function in topological. 
\bibitem{key55} [Nagaveni1] D. Iyappan and N. Nagaveni, On semi generalized b-closed set, Nat. Sem. On Math and Comp. Sci,, (2010) Proc.6.
\bibitem{key56} [Nagaveni5] N. Nagaveni, Studies on generalizations of homeomorphisms in topological spaces, Ph.D., Thesis, Bharathiar University, Coimbatore (1999).
\bibitem{key57} [Nagaveni2] A. Narmadha, N. Nagaveni and T. Noiri, On Regular b-Open Sets in Topological Spaces Int. Journal of Math. Analysis, Vol. 7, 2013, no. 19, 937--948.
\bibitem{key58} [Nagaveni3] C. Mukundhan and N. Nagaveni, A Weaker Form of a Generalized Closed Set, Int. J. Contemp. Math. Sciences, Vol. 6, 2011, no. 20, 949--961.
\bibitem{key59} [Nagaveni6] C. Mukundhan and N. Nagaveni, On semi weakly g homeomorphisms and semi weakly g irresolute maps in topological spaces, Advances in Theoretical and Applied Mathematics, Volume 6, No. 5 (2011), 591--601.
\bibitem{key60} [Nagaveni7] L.Vinayagamoorthi, N. Nagaveni, On Generalized-$\alpha$ b closed sets in topological spaces bitopological spaces and fuzzy topological spaces, Ph.D Thesis, Anna University, Tamil nadu, India (2014).
\bibitem{key61} [Navalagi2] Govindappa Navalagi, S.V. Gurushantanavar, and Chandrashekarappa A.S,  On $\alpha$ Generalized Preclosed Sets In Topological Spaces, International Journal of Mathematics and Computing Applications, Vol. 2, Nos. 1--2, January-December 2010, pp. 35--44.
\bibitem{key62} [Navalagi3] G B Navalagi, On Semi pre Continuous Functions and Properties of Generalized semi pre closed in Topology, IJMS 2002, 29(2), 85--98.
\bibitem{key63} [Navalagi] Govindappa Navalagi and Chandrashekarappa, On gspr closed sets in topological spaces, Vol.2, Nos. 1--2, (2010), pp.51.-58.
\bibitem{key64} [Njastad] O. Njastad, On some classes of nearly open sets,  Pacific J. Math., 15(1965), 961--970.
\bibitem{key65} [Noiri2] G. Di Maio and T. Noiri, On s-closed spaces, Indian J. Pure Appl. Math., 18(1987), 226--233.
\bibitem{key66} [Popa] T. Noiri and V. Popa, On g-regular spaces and some functions, Mem. Fac. Sci. Kochi Univ. Math 20 (1999) 67--74.
\bibitem{key67} [Noiri] T. Noiri, A generalization of closed mappings, atti. Acad. Naz.lincei rend, cl, Scifis. Mat. Natur. 1973; 54:412--415.
\bibitem{key68} [Noiri1] T. Noiri, H. Maki and J. Umehara, Generalized preclosed functions, Mem. Fac. Kochi Univ. Ser. A, Math., 19(1998), 13--18.
\bibitem{key69} [Palaniappan] N. Palaniappan and K.C. Rao, Regular generalized closed sets, Kyungpook, Math. J., 33 (1993), 211--219.
\bibitem{key70} [Palaniappan1] Y. Palaniappan, On Regular Generalised B-Closed Sets,  International Journal of 	Scientific \& Engineering Research, Volume 4, Issue 4, April-2013 14--10.
\bibitem{key71} [Parimelazhagan] R. Parimelazhagan and V. Subramonia Pillai, ``Strongly g* closed sets in topological spaces'',	Int. Journal of Math. Analysis, Vol. 6, 2012, no. 30, 1481--1489.
\bibitem{key72} [Park]H. P. Park and J. K. Park, On Semi-generalized locally closed sets and SGLC-continuous functions, Indian J, Pure Appl. Math., 31(9) (2000), 1103--1112.
\bibitem{key73} [Patty] C. W. Patty, Bitopological spaces, Duke Math. J., 34 (1967), 387--392.
\bibitem{key74} [Pious] C. Devamanoharan, S. Pious Missier and S. Jafari, -Closed sets in Topological Spaces,
\bibitem{key75} [Pious2] C. Deva manoharan, S.Pious Missier and S. Jafari, European Journal Of Pure And	Applied Mathematics, Vol. 5, No. 4, 2012, 554--566.
\bibitem{key76} [Pious1] N. Palaniappan, J. Antony Rex Rodrigo and S. Pious Misssier, ``On Closed sets in Topological Spaces'', Varahamihir Journal of Mathematical Sciences, Vol, 7 No 1 (2007) 33--38.
\bibitem{key77} [Pious]	S. Pious Missier, K. Ali, and A. Subramanian, ``g\#p  Closed Sets in Topological Spaces'', International Journal of Mathematical Archive, 4 (1) (2013) 176--181.
\bibitem{key78} [Popa] V. Popa, On certain properties of bitopological semi-separate spaces, Mat. Vensnik., 3(1979), 71--75.
\bibitem{key79} [Pushpalatha] A. Pushpalatha, Studies on generalizations of mapping in topological spaces, Ph.D Thesis, Bharathiar university (2000), Coimbatore.
\bibitem{key80} [Rajamani] Rajamani M., Viswanathan K, On $\alpha$gs-closed sets in topological spaces, Acta Ciencia Indica,
\bibitem{key81} [Reilly] I. L. Reilly, On bitopological separation properties, Nanta Math., 29(1972), 14--25.
\bibitem{key82} [Reilly1] M. Ganster and I. L. Reilly, Locally closed sets and LC-continuous functions Internat. J. Math. Sci., 12 (1989), 417--424.
\bibitem{key83} [SampathKumar] S. SampathKumar, $\beta$-open sets, $\beta$-continuity and $\beta$-open mappings in bitopological spaces, Journal of the Indian Math. Soc., 62(1996), 187--195.
\bibitem{key84} [Sanjay] Sanjay Mishra, Nitin Bhardwaj and Varun Joshi, On Regular Generalized weakly (rgw)-Closed Sets in Topological Spaces, Int. Journal of Math. Analysis, Vol. 6, 2012, no. 39, 1939 1952.
\bibitem{key85} [Sanjay1] Sanjay Mishra, Nitin Bhardwaj and Varun Joshi, On Generalized Pre Regular Weakly (gprw)-Closed Sets in Topological Spaces, International Mathematical Forum, Vol. 7, 2012, no. 40, 1981 1992.
\bibitem{key86} [Sheik1] M. Sheik John, A study on generalization of closed sets on continuous maps in topological and bitopological spaces, Ph.D Thesis, Bharathiar University, Coim-batore, (2002).
\bibitem{key87} [Shyala] Shyala Isac Mary T, P Thangavelu.on Regular pre-semi closed sets in topological spaces, KBM J of Math Sc \& comp Applications. 2010;1:9--17.
\bibitem{key88} [Stone] M. Stone, Application of the theory of Boolean rings to general topology, Trans. Amer. Math Soci  41 (1937), 374--481.
\bibitem{key89} [Subashini1] Subashini jesu rajan, On $\beta$wg set and Continuity in Topological Spaces, International journal of computing, Vol 4 Issue 3, July 2014.
\bibitem{key90} [Subashini] Subashini jesu rajan, On $\beta$wg** set and Continuity in Topological Spaces, International journal of  computing, Vol 4 Issue 3, July 2014.
\bibitem{key91} [Sheik] M. Sheik John, On w-closed sets in Topology, Acta Ciencia Indica, 4 (2000), 389--392.
\bibitem{key92} [Sundaram] P. Sundaram, Studies on Generalizations of Continuous maps in Topological Spaces, Ph.D. Thesies, Bharathiar University, Coimbatore, 1991.
\bibitem{key93} [Shaik John] T. Fukutake, P. Sundaram and M. Shaik John, w-closed sets, w-open sets and w-continuity in bitopological spaces, Bull. Fukuoka Univ. Ed. part III, 51 (2002), 1--9.
\bibitem{key94} [Vadivel] Vadivel A, K. Vairamanickam, rg$\alpha$-Homeomorphisms in Topological Spaces, Int. Journal of Math. Analysis. 4 (2010) 881--890. 
\bibitem{key95} [Veera Kumar3] M.K.R.S. Veera Kumar, On g$\hat{a}$closed sets in topological spaces, Bulletin Allahabad Math. Soc.18(2003),  99--112.
\bibitem{key96} [Veera kumar2] M.K.R.S. Veera Kumar, Between Semi Closed sets and Semi Pre Closed sets, Rend.Istit.Mat. Univ. Trieste. Vol. XXXI, 25--41 (2000).
\bibitem{key97} [Veera kumar] M.K.R.S. Veera kumar, g\#-closed sets in topological spaces, Mem. Fac. Sci. Kochi Univ. Ser. A. Math 24(2003), 1--13. 
\bibitem{key98} [Veera Kumar1] Veerakumar M.K.R.S, g\# semi-closed sets in topological spaces. International Journal of Scientific \& Research Publications, Vol 2,Issue 6, June 2012 ISSN 2250--3153.
\bibitem{key99}  [Velicko] N. V. Velicko, H-closed topological spaces. Amer. Math. Soc. Transl. 78 (1968), 103--118.
\bibitem{key100} [Vigneshwaran1] M. Vigneshwaran and A. Singaravelan, Applications of {*}{*}g$\alpha$-closed sets in Topo-logical Spaces, IJMA-5(10), 2014, 139--150.
\bibitem{key101} [Vigneshwaran] M. Vigneshwaran and R. Devi, On $G \alpha O$-kernel in the digital plane, International Journal of Mathematical Archive-3(6), 2012, 2358--2373.
\bibitem{key102} [Wali2] R. S. Wali and Prabhavati S. Mandalgeri, On $\alpha rw$-Closed and $\alpha rw$-Open maps Maps in Topological Spaces, International Journal of Applied Research, 1(11) 2015, 511--518.
\bibitem{key103} [Wali3] R. S. Wali and Prabhavati S. Mandalgeri, on $\alpha rw$-closed sets in topological spaces. Int. Journal of Mathematical Archive-5(10), 2014, 68--76.
\bibitem{key104} [Wali4] R. S. Wali and Prabhavati S. Mandalgeri, On $\alpha rw$-Continuous and $\alpha rw$-Irresolute Maps in Topological Spaces, IOSR Journal of MathematicsVol. 10, Issue 6, Ver. Vi Dec. 2014, 14--24.
\end{thebibliography}
