\chapter{Gem Of A Teacher}

\begin{center}
\Authorline{Srm.Chitra. k. Viswas}

\smallskip

Cycle Pure Agarabattis\\
Project lead
\addrule
\end{center}
Shri Gangadhar Bhat is my spiritual guide and teacher for at the age of 58 years I learnt the shlokas, “Vakratunda Mahakaya….” \& “Saraswati Namastubhyam...........”.and meaning from him !

I approached him on behalf of Cycle Pure Agarbathies with the idea of making a Do-it-yourself prayer assortment for those (like me) who do not know, but would like the next generation in the family to know and to keep in touch with our glorious Indian culture and tradition.

There is so much one can learn just listening to Gangadhar Bhat Sir. I have personally imbibed a whole lot more than my work at Cycle Pure Agarbathies required. His thoughts are so evolved that he never imposes his beliefs. On the contrary,this gentleman teacher is also happy to learn if he must.

Despite being a traditionalist he taught and guided me well enough to collate some Hindu prayer instructions. His contributions significant in preserving our prayer rituals and tradition, and+ appropriately guiding the emerging generations world over.

Though Gangadhar Sir is a dedicated teacher at heart and not a businessman, he understood the nuances of manufacturing business. He would patiently explain to the smallest detail, how a prayer had to be done and sometimes mention that this is not a universally feasible path.Sure enough, a couple of days later, after discussions with peers and reading different scriptures, he would come up with an apt pan-India solution.

I have been bothering him for information both at work and at home for many years. He is a very pleasant-natured person who never says “no” to anyone who wishes to learn. Another very important point to mention here is that all this has been possible because of his wife Shailaja’s smiling, welcoming nature. She is the most co-operative back office computer support too. 

I understand Gangadhar Bhat Sir’s need and passion to move back to his village and take on family responsibilities of another kind now that he has retired from College teaching. But, Mysore city will lose a very valuable asset if he does so. However, I am sure he will be available to everyone long distance and his front door at home wherever he will decide to reside, will also always be open to seekers of information and knowledge.

On behalf of Cycle Pure Agarbathies,I take this opportunity to sincerely say “Thank you Gangadhar Bhat Sir.We wish you and your wife Smt. Shailaja, good health, peace and happiness always”.

\articleend
