\chapter{Signatories To Letter Shared With California School Districts}

\centerline{\underline{27 US based faculty}} 

\begin{multicols}{2}
\begin{trivlist}
\itemsep=10pt
\item \textbf{Kundan Singh, Ph.D.}\\ 
Former Executive Core Faculty\\
Adjunct Research Faculty\\
Sofia University. Palo Alto. CA\\
Adjunct Faculty, California Institute of Integral Studies.\\ 
San Francisco. CA. 

\item \textbf{Ashok Aklujkar, Ph.D.}\\ 
Professor Emeritus and Former Head,\\ 
Department of Asian Studies,\\ 
University of British Columbia.\\ 
Canada.

\item \textbf{June R. Klein, EdD, MBA, CPA}\\ 
Professor, Palo Alto University.


\item \textbf{Ronald L. F. Davis, Ph.D.}\\ 
Professor Emeritus\\ 
California State\\ 
University-Northridge. CA.

\item \textbf{Madan L. Goel, Ph.D.}\\ 
Professor Emeritus of Political Science\\
University of West Florida\\
Pensacola, Fl 32514.

\item \textbf{Subhash Kak, Ph.D.}\\ 
Professor\\
Oklahoma State University\\
Author of 10 + books on Ancient India.\\

\item \textbf{Doug Oman, Ph.D.}\\
Associate Adj. Professor\\
University of California, Berkeley\\
School of Public Health\\
Past-president, Society for the\\ Psychology of Religion\\ 
and Spirituality (Division 36 of\\ 
the American Psychological\\ 
Association;\\  
for identification purposes only.) 
 
\item \textbf{Nilesh Oak}\\ 
Adj. Professor\\
School of Indic Studies\\
University of Massachusetts\\ Dartmouth.
 
\item \textbf{Bal Ram Singh, Ph.D.}\\ 
President, Institute of Advanced\\ Sciences\\[2pt]
Professor and Director,\\ 
Botulinum Research Center\\[2pt]
Professor and Executive Mentor,\\ 
School of Indic Studies\\
Institute of Advanced Sciences\\
Dartmouth, MA 02747\\[2pt]
Former Tenured Professor,\\ 
University of Massachusetts\\ Dartmouth\\[2pt]
Former Visiting Professor,\\ 
Jawaharlal Nehru University,\\ 
New Delhi\\[2pt]
Former Visiting Professor,\\ 
Harvard Medical School\\[2pt]
Former Visiting Professor,\\ 
Georgetown University\\[2pt]
Former Visiting Professor,\\ 
Yang Ming University,\\ 
Taiwan.
 
\item \textbf{Annu Maheshwari}\\ 
Former Faculty\\
Brandeis University.
 
\item \textbf{James T. Richardson, Ph.D., J.D.}\\ 
Foundation Emeritus Professor of\\ Sociology\\ 
and Judicial Studies\\
University of Nevada, Reno.
 
\item \textbf{T.K. Roy, Ph.D.}\\ 
Prof. Emeritus, Indic Studies\\
University of Massachusetts\\
Dartmouth.
 
\item \textbf{Michael K. Ward, Ph.D.} \\
Associate Professor, History\\
Department Chair, Social Sciences\\
Ventura College.\\[2pt]
Visiting Lecturer in History at\\ California\\
State University, Northridge. 

\item \textbf{Arvind Sharma, Ph.D.}\\ 
Birks Professor of Comparative\\ Religion\\ 
McGill University. Canada.
 
\item \textbf{Barbara A. McGraw, J.D., Ph.D.}\\ 
Professor, Social Ethics, Law, and Public Life,\\[2pt]
Director, Center for Engaged\\ Religious Pluralism\\[2pt]
Director, Interfaith Leadership\\ Program
Saint Mary’s College of\\ California. CA.
 

\item \textbf{Michael Nagler, Ph.D.}\\
Prof. emer., UC, Berkeley\\
The Metta Center for Nonviolence,\\ 
Founder and President\\
Nonviolence Radio, co-host\\ 
Daily Metta, editor.
 
\item \textbf{Lavanya Vemsani, Ph.D.}\\ 
Ph.D (History) Ph.D.\\ (Religious Studies)\\
Professor, Dept.\ of Social Sciences\\
Shawnee State University. OH.\\ 
Co-founder, American Academy of\\ Indic Studies\\ 
Editor-in-Chief\\ 
American Journal of Indic Studies.
 
\item \textbf{Pankaj Jain, Ph.D.}\\ 
Prof Philosophy \& Religion\\
Co-Chair, India Initiative Group\\
University of North Texas\\[2pt]
Co-founder, American Academy of\\ Indic Studies\\[2pt]
Exec.\ Editor, American Journal of\\ Indic Studies\\
Section Editor for Hinduism,\\[2pt] 
Encyclopedia of Indian Religions.
 
\item \textbf{Mona Khaitan, Ph.D.}\\ 
Professor, Hindu Studies\\
MassBay Community College.
 
\item \textbf{P. Scott Corbett, Ph.D.}\\ 
Professor, History, Ventura College,\\ 
and Lecturer, History,\\ 
California State University at\\ 
Channel Islands. CA. 
 
\item \textbf{Nalini Rao, Ph.D.}\\ 
Assoc. Prof. World Art\\
Soka University\\
Ph.D. (Ancient History \& Archelogy),\\ Ph.D (Art history).
\vfill
 
\item \textbf{Jeffery D. Long, Ph.D.}\\ 
Professor of Religion and\\ Asian Studies\\
Elizabethtown College. PA.
 
\item \textbf{Sabina Magliocco, Ph.D.}\\ 
Professor, Department of\\ Anthropology\\
California State University.\\ Northridge. CA. 
 
\item \textbf{Lynn C. Waelde, Ph.D.}\\ 
Professor\\
Director, Meditation and Psychology Area of Emphasis,\\[2pt] 
Clinical Psychology Ph.D. Program\\[2pt]
Director and Founder,\\ Inner Resources Center\\[2pt]
Palo Alto University.
 
\item \textbf{Ramesh N. Rao, Ph.D.}\\ 
Professor and Associate Chair,\\
Department of Communication\\
Columbus State University.
 
\item \textbf{Anuja Kabra}\\ 
Faculty\\
Rutgers University.
 
\item \textbf{B.V.K. Sastry, Ph.D.}\\ 
Professor, Yoga Samskrutham University,\\ 
Florida. 
\end{trivlist}
\end{multicols}

\newpage

\centerline{\underline{9 India based faculty}} 

\begin{multicols}{2}
\begin{trivlist}
\itemsep=10pt
\item \textbf{Michel Danino}\\ 
Guest Professor,\\ 
Indian Institute of Technology,\\ 
Gandhinagar. India.\\
Member, Indian Council of Historical Research,\\ 
New Delhi. India.

\item \textbf{Meenakshi Jain, Ph.D.}\\ 
Member, Governing council Indian Council of Historical Research,\\
India.
 

\item \textbf{Madhu Purnima Kishwar}\\ 
Maulana Azad National Professor\\
India Council of Social Research\\
Founder Manushi. India.
 
\item \textbf{Alex Hankey, Ph.D.}\\ 
Distinguished Professor of Yoga\\
Swami Vivekananda Yoga\\ Anusandhana Samsthana.\\ 
Bangalore. India.
 
\item \textbf{Neelam Luthra, Ph.D.}\\ 
Associate professor\\
Sohan Lal DAV college of Education\\
Ambala City. India.
 
\item \textbf{Sharad Joshi, Ph.D.}\\ 
Former Pro Vice-Chancellor\\
M. S. University. Baroda. India.

\item \textbf{B. B. Lal, Ph.D.}\\ 
Former Director General,\\ 
Archeological Survey of India,\\
Author of 27 books on Ancient Indian History (9 of which are on Indus Sarasvati and Aryan people) and 150+ research papers and articles in scientific journals. 
 
\item\textbf{Choodamani Nandagopal, Ph. D.}\\ 
(History \& Archaeology)\\[2pt]
Art Historian, Author,\\[2pt]
UNESCO Fellow\\[2pt]
Former Dean, School of Humanities \& Social Sciences,\\ 
Jain University, Bangalore.\\[2pt]
Member Advisory Academic\\ Committee,\\ 
IGNCA, Bangalore.\\[2pt]
Author: Arts and Crafts of Indus\\ Civilisation,\\ 
Temple Treasures 3 volumes,\\[2pt]
Dance and Music in Temple\\ Architecture,\\ 
Ritual Enactment in Temple\\ Tradition.
 

\item \textbf{Ambika Datta Sharma, Ph.D.}\\ 
Professor\\ 
Department of Philosophy\\
Dr. Hari Sing Gour Univrsiy.\\ 
India.
\end{trivlist}
\end{multicols}
\newpage

\centerline{\underline{3 Faculty from other countries}}

\begin{multicols}{2}
\begin{trivlist}
\itemsep=12pt
\item \textbf{Anand Paranjpe, Ph.D.}\\ 
Professor Emeritus, Simon\\ Fraser University.\\ 
Vancouver, Britsh Columbia.\\ 
Canada.
 
\item \textbf{T. Ramakrishna, Ph.D.}\\ 
Formerly, Senior Visiting Fellow\\
Exeter College, University of Oxford, UK.
\vfill

\item \textbf{Prof. (FH) Regina Rowland, BA, MGD, PhD}\\ 
Departments of Business Studies,\\ 
Energy \& Environmental\\ Management\\
University of Applied Sciences\\ Burgenland, Austria\\
Campus 1, A-7000 Eisenstadt\\
Steinamangerstr. 21, A-7423\\ Pinkafeld. Austria.\\
Former Professor of Design\\ Management,\\ 
Savannah College of Art and Design,\\ Savannah. GA.
\end{trivlist}
\end{multicols}
