\chapter{Appendix to Part II: List of Publications}

\begin{enumerate}
\itemsep=0pt
\item $\wedge$ - binding in hypernuclei by nonlocal interaction, (with
S.N. Biswas),  Phys. Rev. {\bf 122}, 712 (1961). 

\item The spins and lifetimes of the light hypernuclei, (with R.H.
Dalitz), Phys. Letters {1}, 58 (1962).

\item Resonance poles and mass differences within unitary multiplets, 
(with R.H. Dalitz), Phys. Letters {\bf 7}, 373 (1963).

\item The binding of $\wedge \wedge$ - hypernuclei, (with R.H. Dalitz), 
Nucl. Phys. {\bf 50}, 450 (1964).

\item Scattering amplitudes on unphysical sheets and resonance poles, 
Nuovo Cimento {\bf 31}, 697 (1964).

\item Meson-Baryon mass splittings and resonance multiplets in $SU_3$
symmetry, Nuovo Cimento {\bf 37}, 1004 (1964).

\item A model calculation for the $Y^\star_o$ (1405) Resonant State, 
(with R.H. Dalitz and T.C. Wong), Phys. Rev. 153, 1617 (1966).

\item Decimet Baryons as ``Elementary Particles",\ Phys.\ Rev.\ {\bf
159}, 1488 (1967).

\item Current Commutator and CPT,\ Phys.\ Rev.\ {\bf 160}, 14 27 (1967).

\item $Y^\star_o$ (1405) as a possible exception to the quark-picture of
Hadrons - {\it Proc. of the Tenth Symposium on Cosmic Rays, Elementary
Particles and Astrophysics}, Aligarh, 1967, P.521.

\item Radiative Corrections to $\beta$ decay - {\it Proc. of the Tenth\break
Symposium on Cosmic Rays, Elementary Particles and\break Astrophysics}, 
Aligarh, 1967, P.507.

\item Does the electromagnetic current have an I = 2 component? (with
P.P. Divakaran and V. Gupta), Phys. Rev. 166, 1792 (1968).

\item Can $Y^\star_o$ (1405) be a bound state of three quarks? Symposia
on Theoretical Physics and Mathematics (Plenum Press), Vol. {\bf 9}, 531
(1969).

\item $SU(3)$ Symmetry Breaking in Semileptonic Decays, (with L.K.
Pandit), Nucl. Phys. {\bf B 9}, 531 (1969).

\item Restrictions on Schwinger Terms due to Lorentz Covariance, (with
V. Gupta), Nucl. Phys. {\bf B 10}, 11 (1969).

\item Sum Rules from Local Current Algebra, (with V. Gupta), Phys. Rev.
{\bf 185}, 1940 (1969).

\item On the Electron-Muon Universality in Strangeness\break Changing
Processess, (with K.V.L. Sarma), Nucl. Phys. {\bf B 18}, 568 (1970).

\item The Leptonic K-Decays with Broken $SU(3) \times SU(3) \mbox{and}
\ SU(3)$ and Violation of $(\mu - e)$ - Universality, (with S.C. Chhajlany
and L.K.Pandit), Nucl. Phys. {\bf B 21}, 1 (1970).

\item Axial Vector $K_{\ell 4}$ - Decay From-factors based on Current
Algebra, (with S.C. Chhajlany and L.K. Pandit), Phys. Rev. D2, 1934
(1971).

\item Inelastic Electron-Proton Scattering and a Sum Rule for the
Schwinger Term,\ (with V.\ Gupta),\ Phys.\ Rev.\ D3,\ 677 (1971).

\item A fixed pole in the Virtual Compton Amplitude $A_2$, (with R.
Rajaraman), Phys. Rev. {\bf D3}, 266 (1971).

\item Schwinger Terms, Fixed Poles and Inelastic $e-p$ Scattering, {\it Proc.
of Symposium on Particles and Fields}, University of Madras, 1971, p.113.

\item Effective-Dipole Form-Factors for the $K_{\ell_3}$ - Decay, (with
S.C. Chhajlany and L.K. Pandit), Phys. Letters 35B, 166 (1971).

\item Empirical Test for Composite Hadrons, Phys. Rev. {\bf D5}, 610
(1972).

\item Yang-Mills Fields and Theory of Weak Interactions, Report of Saha
Institute Lectures (1971).

\item Divergences of the Higher Order Corrections to $\mu$- Decay in the
Gauge Theory, Phys. Rev. {\bf D6}, 3032 (1972).

\item Scaling and Asymptotic Behaviour of Form Factors, (with P.P.
Divakaran), Nucl. Phys. {\bf B 60}, 437 (1973).

\item Contribution of Neutral Currents to the Decay $K_L \rightarrow
\mu^+ \mu^-$ (with S.H. Patil), Unpublished report TIFR/TH/72-37.

\item Unification of weak and Electromagnetic Interactions, {\it Proc. of the
First Symposium on High Energy Physics, Bombay}, (Dept. of Atomic
Energy, 1972), p.331.

\item Analysis of the Neutral-Current Interaction in the Inclusive
Neutrino Reactions, (with K.V.L. Sarma), Pramana {\bf 2}, 62 (1974); E 225.

\item Gauge Theories and the Neutral-Current --- New Developments in Weak
Interactions,  Physics News {\bf 5}, 6 (1974).

\item Model-Independent Analysis of the Neutral-current Interac\-tion in
the Inclusive Neutrino Reactions, (with K.V.L.\break Sarma), Pramana {\bf 3}, 
44 (1974).

\item Bounds on the Coupling Constants of the Neutral Current, (with
K.V.L. Sarma), Physics Letters {\bf 55 B}, 201 (1975).

\item Theoretical Implications of the Resonance Anomalies in $e^+ e^-$
System, (with S. Pakvasa and S.F. Tuan), Phys. Rev. {\bf D 11}, 1345 (1975).

\item General Space-Time Structure of the Neutral-Current\break Interactions,\ 
(with S.\ Pakvasa),\ Phys. Rev.\ {\bf D 12},\ 113 (1975).

\item Comments on ``Is Bound Charm Found"? (with J. Pasu\-pathy), 
Phys. Rev. Lettt. {\bf 34}, 1250 (1975).

\item Effect of Gluons on Neutral-Current Interactions, (with\break Probir
Roy), Pramana {\bf 4}, 222 (1975).

\item Interpretation of the new particle of the cosmic ray neutrino
experiment, (with K.V.L. Sarma), Pramana {\bf 5}, 78 (1975).

\item Charmed Particles and Anomalies in Antineutrino Collisions, (with
K.V.L. Sarma), unpublished report. 

\item Colour gluons and scaling in a Unified Gauge Model, (with Probir
Roy), Pramana {\bf 5}, 303 (1975).

\item Radiative Decays, Photoproduction and Total Cross Section of
$\psi$ - particles, (with K.V.L. Sarma and J. Pasupathy), Pramana {\bf
6}, 29 (1976).

\item Colour Gluon Excitation at Fermilab, (with Probir Roy), Phys. Rev.
Lett. {\bf 36} 355, 689 E (1976).

\item Integral versus Fractional Charges of Quarks - A view\break  through Gauge Theory, {\em Few body Dynamics}, Edited by A.N. Mitra etal, North
Holland (1976), p. 341.

\item General Bounds on the Isovector coupling constants of the weak
neutral current, (with K.V.L. Sarma), Pramana {\bf 7}, 194 (1976).

\item Field Theory of Extended objects (Invited talk), {\it Procee\-dings of
the Third Symposium on High Energy Physics,\break Bhubaneswar} (Dept. of
Atomic Energy, 1976), Vol.II p.163.

\item Renormalizable Fermi Theory of Weak Interactions (Invited Talk), 
{\it Proceedings of the International symposium on\break Frontiers of Theoretical
Physics} (University of Delhi, 1977), p.121.

\item Renormalizable Theories from Nonrenormalizable Interactions (with
V. Srinivasan), Pramana {\bf 9}, 239 (1977).

\item Generation of Gluons from Quark Confinement (with V. Srinivasan), 
Pramana, {\bf 10}, 33 (1977).

\item Solitons, Monopoles and Bags (Invited talk), {\it Proceedings of the
Nuclear Physics and Solid State Physics Symposium, Poona} (Dept. of
Atomic Energy, 1977) p. 1.

\item Gauge Theory, {\it Proceedings of the Anniversary Symposium of
Matscience, Madras} (1978).

\item Recent Developments in Gauge Theory, {\it Advances in High Energy
Physics} (Tata Institute of Fundamental Research, Bombay), 
Vol. {\bf 3}, 333 (1977).

\item Neutral Current Cross sections for Neutrinos on Protons (with
K.V.L. Sarma) - Pramana {\bf 11}, 609 (1978).

\item Classical Solutions to Quark Confinement (with V. Srinivasan), 
Pramana {\bf 11}, 739 (78).

\item The possibility of two-gluon jets in electron-positron
annihilation (with S.D. Rindani), Phys. Letts. {\bf 83 B}, 107 (1979).

\item Neutral Currents in left-right symmetric gauge models, (with J.K.
Bajaj), Pramana {\bf 12}, 397 (1979).

\item General relations among observables in neutral-current\break phenomena
mediated by one or two Z bosons, (with M.K. Parida), Pramana {\bf 13}, 
183 (1979).

\item Linking weak interactions with electrodynamics (Popular\break article)
Science Today, January 1979, p 17.

\item The electroweak mixing angle in unified gauge theories (with J.K.
Bajaj) Pramana {\bf 14}, 395 (1980).

\item The electroweak mixing angle in unified gauge theories - the
renormalization effects (with J.K. Bajaj), Pramana {\bf 14}, 411
(1980).

\item Constraints on the parameters of the Left-Right symme\-tric gauge
models (with J.K. Bajaj) Phys. Lett. {\bf 93 B}, 461 (1980).

\item Electron-positron annihilation into three jets in the
integer-charge quark model (with S.D. Rindani), Phys. Lett. {\bf 94 B}, 
76 (1980).

\item Gauge Theories - A Review, {\it Proceedings of the Einstein Centenary
Symposium}, Ahmedabad, 1979, p. 208.

\item Spin-dependance of the charged gluon contribution in deep
inelastic scattering (with H.S. Mani and S.D. Rindani),\break Physics Lett.
95B, 107 (1980).

\item Contribution of coloured Higgs bosons to $e^+ e^-$ annihilation
into two jets (with S.D. Rindani), Physics Lett. {\bf 99 B}, 361 (1981).

\item Contribution of gluons and coloured Higgs bosons to $e^+ e^-$
annihilation into three jets (with S.Lakshmi Bala and S.D. Rindani), 
Phys. Lett. {\bf 105 B}, 477 (1981).

\item Integer-charged quark model and electorn-positron annihilation
into three jets (with S.D. Rindani), Prog. Theor. Phys. {\bf 67}, 1505 (1982).

\item Contributions of neutral current couplings in arbitrary\break two-Z
boson extensions of the Weinberg-Salam Model (with J.K. Bajaj)
Unpublished report.

\item BAN in unorthodox models - Invited talk given at the\break {\it International
Colloquium on Baryon Nonconservation} held at Bombay (January 1982)
(published in the Proceedings, p.65).

\item Large-N gauge theory of loops and strings I and II (with H. Bohr)
Phys. Rev. {\bf D 32}, 1547, 1553 (1985).

\item Do the PETRA two-photon jet experiments rule out\break integrally
charged quarks?  (with T. Jayaraman and S.D. Rindani), Phys. Lett. {\bf
119 B}, 215 (1982).

\item A note on the annihilation of polarised $e^+ e^-$ beams into jets
in the broken colour gauge theory (with S. Lakshmi Bala and S.D.
Rindani), Phys. Lett. {\bf 117B}, 345 (1982).

\item Nonrenormalization of the weak vertex in gauge theories with
integrally charged quarks (with M.S. Sriram), Pramana {\bf 19} 315
(1982). 

\item The mass of the electron neutrino and electron capture in
$^{163}Ho$ (with S. Yasumi etal), Phys. Lett. {\bf 122 B}, 461 (1983).

\item Validity of the equivalent-photon approximation for the\break production
of massive spin-l particles (with T. Jayaraman and S.D. Rindani), Pramana
{\bf 26}, 21 (1986).

\item Is colour broken by a monopole? (with R. Ramachandran), 
unpublished report.

\item Grand Unified Theories - Invited Talk at the {\it VI High\break Energy
Physics Symposium, Mysore}, (Dept. of Atomic\break Energy, 1982) (Proceedings p.1).

\item Magnetic Monopole - Invited Talk at {\it Thirty Years\break
Commemoration symposium}, Saha Institute of Nuclear\break Physics, Calcutta, 
1983 (Proceedings p.87).

\item Decays of W and Z in the broken-colour model (with S. Lakshmibala
and S.D. Rindani), Pramana {\bf 24}, 423 (1985).

\item The recent single-tag two-photon experiments at PETRA and the
issue of quark charges (with R.M. Godbole, J.C. Pati, S.D. Rindani and
T. Jayaraman), Phys. Lett. {\bf 142 B}, 91 (1984).

\item Fields with vanishing colour-currents (with V. Srinivasan and M.S.
Sri Ram), Pramana {\bf 24}, 435 (1985).
 
\item Modification of the equivalent-photon approximation for the
production of massive spin-1 particles (with T. Jayaraman and S.D.
Rindani), (unpublished report).

\item Direct Photon Production in $e^+ e^-$ collisions (with T.
Jaya\-raman and S.D. Rindani) - Phys. Rev. {\bf D 32}, 1262 (1986).

\item Two-photon production in $pp$ and $p\bar{p}$ collisions (with T.
Jaya\-raman and S.D.\ Rindani) - Phys.\ Rev.\ {\bf D 33},\ 672 (1986).

\item What is the next step after electroweak unification? in {\em
Recent Advances in Theoretical Physics}, (World Scientific 1985) Ed: R.
Ramachandran, p 89-144.

\item Quantum Interference Effects in a five-dimensional Kaluza-Klein
Theory (with R. Parthasarathy and R. Vasudevan) - Classical and Quantum
Gravity {\bf 3}, 425 (1986).

\item $p\bar{p} \rightarrow W^- \gamma X$ as a test of the quark
charges (with S.\ Lakshmi\-bala, X.-G.He and S. Pakvasa), Modern Phys.
Lett. {\bf A} 1, 277 (1986).

\item Higgs Couplings in the integer quark model (with S.\ Lakshmi\-bala), 
Pramana {\bf 27}, 371 (1986).

\item $e^+e^- \rightarrow \gamma + 2$ jets as a test of quark charges
(with X.-G.He, S. Pakvasa and S.D. Rindani), Phys. Lett. {\bf B 185}, 
158 (1987).

\item Quantum \ Interference \ Effects \ in \ a \ generalized \ Kaluza-Klein \
Theory
\ (with \ R. Parthasarathy \ and \ P. Nagarajan) - \ Classical \ and \ Quantum \ Gravity
{\bf 4} 1327 (1987).

\item Perspectives in High Energy Physics - Invited talk, \ in {\it VIII High
Energy Physics Symposium, Calcutta}, 1986 Procee\-dings p.399.

\item Building up the Standard Gauge Model of High Energy Physics, \ in
{\it Gravitation, Gauge Theories and the Early\break Universe}, Ed: B.R. Iyer etal, 
Kluwer Academic publishers 1989, p.185.

\item Superstrings - an elementary review, {\it Proceedings of the
Internatio\-nal Conference on Changing Faces of the Physics of Particles \&
Nuclei}, BHU, Varanasi (1987), p.195.

\item An Electroweak Factorization Theorem (with S. Lakshmi\-bala)
Int. J. Mod. Phys. {\bf A4}, 2977 (1989).

\item Introduction to String Theories, in {\it Gravitation, Quantum
Fields and Superstrings} (World Scientific, Singapore, 1988) P.333-393.

\item Is Quantum Mechanics Forever? TIFR/TH/89-13 (to be published
in {\it Philosophical Foundations of Quantum Theory}, World Scientific, 
Singapore).

\item Interpretation \ of \ the \ recent \ Kolar \ events \ (with
A. S. Joshipura, V. Gupta and K.V.L. Sarma), Pramana {\bf 33}, 639 (1989).

\item Stability of the S matrix pole (with S.S.Vasan), Pramana {\bf
34}, 183 (1990).

\item Mass \ shifts \ for \ the \ weak \ bosons \ in \ the \ integrally \
charged
\ quark \ model (with N.G.Deshpande), Mod. Phys. Lett. {\bf A5}, 301 (1990).

\item Electroweak radiative corrections and the masses of the\break weak
bosons, in {\it Phenomenology of the Standard Model and Beyond} (World
Scientific, Singapore, 1989) p.407.

\item Electroweak Symmetry, in Proceedings of Raman Centenary
Symposium on {\it Waves and Symmetry}, Current Science {\bf 59}, 1080
(1990).

\item Rosenbluth \ scattering \ with \ Pauli \ moment \ for \ the \
electron
\ (with R.Basu and M.V.N.Murthy), J. Phys. G {\bf 17}, 401 (1991).

\item Algebra for fermions with a new exclusion principle (with
A. K. Mishra), Pramana {\bf 36}, 537 (1991).

\item The discovery of Dirac equation and its impact on
present-day Physics, in {\it Dirac and Feynman: Pioneers in Quantum
Mechanics}, Ed. by R.Dutt and A.K. Ray, Wiley Eastern (1993) p.9.

\item The future of high energy physics (Keynote address), in\break
{\it Particle Phenomenology in the 90's}, Ed. by
A.Datta etal, World Scientific (1992), p.1.

\item Two Topics in Electroweak Physics: Radiative Corrections
and Baryon Number Violation, in {\it Particle Phenomenology in the
90's}, Ed. by A. Datta etal, World Scientific (1992), p.404.

\item New forms of quantum statistics (with A.K. Mishra)\break Pramana
{\bf 38}, L411 (1992).

\item Orthosupersymmetric quantum mechanics (with A. Khare and
A.K. Mishra), Int. J. Mod. Phys. {\bf A8} 1245 (1993).

\item Parasupersymmetric quantum mechanics and its infinite\break order
(with A. Khare and A.K. Mishra), Mod. Phys. Lett. {\bf A8} 107 (1993). 

\item Algebras of creation and destruction operators (with
A.K. Mishra), Pramana -- J. Phys. {\bf 40} 149 (1993).

\item Quantum field theory of orthofermions and orthobosons (with
A.K. Mishra), Mod. Phys. Lett. {\bf A7} 3425 (1992).

\item Complex $q$-mutator algebra and fractional statistics
(with A.K. Mishra), Mod. Phys. Lett {\bf A9} 419 (1994).

\item ``Infinite Statistics'' and its two-parameter deformation
(with A.K. Mishra), Phys. Lett. {\bf A138} 210 (1994).

\item An algebra for ordered particles (with A.K.Mishra) --
Phys. Lett. {\bf A203}, 153 (1995).

\item Is there a future for High Energy Physics? in {\it Proc. Int.
Conf. on Non Accelerator Particle Physics}, Bangalore, 1994 Ed: R.
Cowsik, World Scientific p.4.

\item Generalized Fock spaces, new forms of quantum statistics and
their algebras (with A.K.Mishra) Pramana - J. Phys. {\bf 45} 91 (1995).

\item A new mechanism for neutrino masses in supersymmetric SO(10)
grand unification (with K.S.Babu and M.K.Parida) -- preprint IMSc-94/19
(to be published).

\item New forms of quantum statistics and generalized Fock\break spaces, 
in {\it Bose and 20th Century Physics}, Kluwer Academic Publishers (The
Netherlands) 1995, Ed: Partha Ghose and Abhijit Mookerji (in Press).

\item Generalized Fock spaces, and new forms of quantum\break statistics, in
{\it Current Trends  in Mathematics and Physics
(A tribute to Harish Chandra)} Ed: S.D. Adhikari, 
Narosa Publishing House, New Delhi, 1995, p.138. 

\item An unified treatment of quantum interference effects (with R.
Parthasarathy and R. Vasudevan), in {\it Selected Topics in Mathematical
Physics (Prof. R. Vasudevan Memorial\break Volume)}, Ed: R. Sridhar etal, 
Allied Publishers Ltd.,\break Madras, 1995, p 279.

\item Leptons and quarks as composites of orthofermions, in\break {\it
Selected Topics in Mathematical Physics (Prof. R. Vasu\-devan Memorial
Volume)}, Ed: R. Sridhar etal, Allied\break Publishers Ltd., Madras, 1995, 
p.288.

\item Solar \ and \ atmospheric\ neutrino \ oscillations \ with
three \ flavours (with Mohan Narayan, M.V.N. Murthy and S. Uma Sankar)
Phys. Rev. {\bf D53} 2809 (1996).

\item Atmospheric neutrinos with three flavour mixing (with\break Mohan
Narayan and S. Uma Sankar) Phys. Rev. {\bf D56} 437 (1997).

\item Fock space representation of differential calculus on the
noncommutative quantum space, (with A.K. Mishra) 
J. Math. Phys. {\bf 38} 466 (1997).

\item Solar eclipse and neutrinos (with Mohan Narayan and Rahul Sinha), 
Current Science {\bf 70} 848 (1996), {\bf 72} 417 (E) (1997).

\item Time-of-night variation of solar neutrinos (with Mohan\break Narayan and
Rahul Sinha), Mod. Phys. Lett. {\bf A13} 1915 (1998).

\item Solar neutrinos and the eclipse effect (with Mohan Narayan,  Rahul
Sinha and C.P. Burgess), Phys. Rev. {\bf D60}, 073006 (1999).

\item Dynamics in a noncommutative phase space (with R.P.\break Malik and A.K.
Mishra), Int. Journal of Mod. Phys. {\bf A13}, 4759 (1998).

\item Perspectives in High Energy Physics, Indian J. Phys. {\bf 72A}, 
679 (1998).

\item High Energy Physics in the 21st Century - An overview, Current
Science {\bf 71} 109 (1996).

\item Particles and Fields (Status Report) in {\it {Physics in
India}}, Indian National Science Academy, New Delhi, 1996, p.41.

\item Fock space for noncommutative calculus, in {\it Special\break Functions
and Differential Equations} Ed: K. Srinivasa Rao etal, Allied Publishers
(1998), p.280.

\item Three flavour implications of the result of the CHOOZ\break
Collaboration (with M. Narayan and S. Uma Sankar), Phys. Rev. {\bf D58}, 
031301 (1998).

\item One pion events by atmospheric neutrinos (with K.R.S.\break Balaji and
S. Uma Sankar), IMSc-98/06/26.

\item A new mechanism for neutrino mass (with P.P. Divakaran), Mod.
Phys. Lett. {\bf A14}, 913 (1999).

\item Neutrinos from stellar collapse: effects of flavour mixing (with
G. Dutta, D. Indumathi and M.V.N.Murthy), Phys. Rev. {\bf D 61}, 13009
(2000).

\item Constraints on the mixing angles of Majorana neutrinos (with R.
Adhikari), Phys. Rev. {\bf D 61}, 031301 (1999). 

\item Hierarchical four-neutrino oscillations with a decay option (with
E.Ma and I. Stancu), Phys. Rev. {\bf D 61}, 071302 (R) (2000).

\item A new differential calculus on noncommutative spaces (with R.P.
Malik and A.K. Mishra), math-ph/9908001.

\item Phenomenology of neutrino oscillations, hep-ph/0004154

\item Neutrinos from stellar collapse: comparision of the effects of three and four flavour mixing (with G. Dutta, D. Indumathi and M.V.N. Murthy), Phys. Rev. D 62, 093014 (2000).

\item Three inequivalent mass-degenerate Majorana neutrinos and a model of their splitting for neutrinos oscillations (with R. Adhikari and E. Ma), Phys. Lett. B486, 134 (2000).

\item Light sterile neutrinos from large extra dimensions (with E. Ma and U. sarkar) Phys. Lett. B495, 363 (2000).

\item Neutrinos from stellar collapse: comparision of signatures in water and heavy water detectors (with G. dutta, D. Indumathi and M.V.N. Murthy), hep-th/0101093.

\item Generalized Fock Spaces and New Forms of Quantum Stati\-stics (with A.K. Mishra), To appear in AIP Proceedings: SPIN 2000, Ed: R.C. Hilborn and G.M. Tino.

\item Quantum Field Theory for Orthofermions and Orthobosons (with A.K. Mishra), To appear in AIP Proceedings: SPIN 2000, Ed: R.C. Hilborn and G.M. Tino.

\item Softly broken $A_4$ symmetry for nearly degenerate neutrino masses (with Ernest Ma) Phys Rev D 64, 113012 (2001).

\item The light unstable sterile neutrinos (with Ernest Ma) Phys Rev D 64, 117303 (2001).

\item Anomalous magnetic moment of muon and L-violating\break supersymmetric models (with R Adhikari), Arxiv: hep-ph/ 0107279.

\item Double three-fold degeneracies for active and sterile neutrinos (with Ernest Ma), Mod. Phys. Lett. A 16, 2207 (2001).

\item Radiative stability of neutirno mass textures (with M K Parida and C R Das), Pramana 62, 647 (2004).

\item High scale mixing unification and large neutrino mixing\break angles (with RN Mohapatra and M K Parida), Phys. Rev. D 69, 053007 (2004).

\item New ``square root'' model of lepton family cyclic symmetry (with Ernest Ma), Phys Rev, D 68, 071302 (2003).

\item India-based neutirno observatory, Invited talk at NuFact 03, 5th internationa workshop on Neutirno Factories and Superbeams, June 2003, Columbia University, USA (published in the Proceedings of the Workshop).

\item Working group report: Neutrinos and astroparticle physics, (with Srubabati Goswami etal) WHEPP-8, IIT-Mumbai, Jan 2004, Pramana 63, 1301 (2004).

\item An SO(10) GUT with see-saw masses for all fermions (with Bipin R Desai and Utpal Sarkar), Phys. Lett, B 626, 167 (2005).

\item Threhold effects on quasi-degenearate neutirnos with high-scale mixing unification (with R.N. Mohapatra and M.K. Parida), Phys. Rev. D71, 057301 (2005).

\item Large neutirno mixing angles for type-I see-saw mechanism in SO(10) GUT (with Bipin R Desai and Utpal Sarkar), Arxiv-hep/0504066.

\item Radiative magnification of neutrino mxings in split supersymmetry (with R N Mohapatra and M K Parida), Phys Rev D 72, 013002 (2005).

\item Hexagonal SU(3) unification and its manifestation in the TeV sclae (with Quing-Hong Cao, Shao-Long Chen and\break Ernest Ma), Phys Rev D 73, 015009 (2006).

\item See-saw fermion masses in an SO(10) GUT (with Kaushik Bhattacharya, C R Das, Bipin R Desai and Utpal Sarkar), Phys Rev D 74, 015003 (2006).

\item Neutirno oscillation probabilities:  Sensitivity to parameters (with D Indumathi, MVN Murthy and Nita Sinha), Phys Rev D 74, 053004 (2006).

\item From atoms to quarks and beyond: a historical panorama, India in the worlds of Physics: Then and Now, (Vol XIII Part 1, History of Science...) Ed: A N Mitra, Pearson Longman, p361 (2009); Arxiv: Physics/0602131.

\item Subaltern Science in the South (with M S Raghunathan), Science and Modern India: An Institutional History, c. 1784-1947, Vol XV Part 4, History of Science...) Ed: Uma Das Gupta, Pearson longman, p669 (2010).

\item Nuetrino mixngs and leptoic CP violation from CKM matrix and Majorana phases (with S K Agarwalla, M K Parida and R N Mohapatra), Phys Rev D 75, 033007 (2007).

\item Observing the dark scalar doublet and its impact on the Standard Model Higgs Boson at colliders (wtih Quing-Hong Cao and Ernest Ma), Phys Rev D 76 095011 (2007).

\item Moller and Bhabha scattering in the noncommutative\break Standard Model (with PK Das and NG Deshpande), Phys Rev D 77, 035010 (2008).

\item Are neutrinos Majorana particles? Keyonote address at Workshop on Neutrinoless Double Beta Decay (NDBD07) at TIFR, Mumbai, published in the Proceedings of the Workshop (2007); Arxiv: 0803.4387.

\item Neutrinos condenste as origin of dark energy (with Jitesh R Bhatt, Bipin R Desai, Ernest Ma and Utpal Sarkar), Phys Lett B 687, 75 (2010).

\item TeV scale implication of noncommutative space-time in\break laboratory frame with polarized beams (with Sumit K Garg, T Shreecharan, P K Das and N G Deshpande), JHEP\break 1107:024 (2011).

\item Group velecity of neutrino waves (with D Indumathi,\break Romesh K Kaul and MVN Murthy), Phys Lett 709 413 (2012).

\item Profound Truths, Current Science 101, 727 (2011).

\item Cabibbo angle and the universality of weak interactions, Physics News 41, 21 (2011).

\item A stable masssive charged particle, Arxiv: 1105.5213 (2011).

\item Manpower for fundamental physics expriments, Current\break Science 103, 19 (2012).

\item High energy physics in 2012, Current Science 103, 250 (2012).

\item An angle to tackle the neutrinos, Current Science 103, 622 (2012).

\item Standard Model, Higgs boson and what next?, Resonance 17 956 (2012).

\item Anomalous Kolar events revisited; Dark matter? (with MVN Murthy), Pramana, J. Phys. 82, L 609 (2014).

\item Models for light unstable dark mattter particles (with Ernest Ma and MVN Murthy), under preparation (2014)

\item Anomalous Kolar events: A proposal for a DM decay\break detector at INO (with Vivek Datar, D Indumathi, MVN Murthy and B Satyanarayana), underpreparation (2014).

\item Predictions from high-sclae mixing unification hypothesis (with Gauhar Abbas, Saurabh Gupta and Rahul Srivastava), Phys Rev D 89, 093007 (2014).

\item High scale mixing unification for Dirac neutrinos (with\break Gauhar Abbas, Suarabh Gupta and Rahul Srivastava), Arxiv: 1312.7384 (2013).

\item Fermi and the theory of weak interactions, Resonance 19, 18 (2014)

\item Physics potential of the ICAL detector of INO (many\break authors), arxiv:1505.07380.

\item High scale mixing relation as a neutral explanation for\break large neutrino mixing (with Gauhar Abbas, MZ Abyaneh, Saurabh Gupta, Monalisa Patre and R Srivastava) In. J. of Mod. Phys A 31-1650095 (2016).

\item Stable or unstable light Dark matter (with Ernest Ma and MVN Murthy) Arxiv: 1507.07609.

\item The story of the neutrinos, arxiv: 1606.08715.

\item Cobimaximal neutirno mixing from $A_4$ and its possible\break deviation (with Ernest Ma), arxiv: 1708.02208.

\item Anomalous Kolar events and Dark Matter Decay in\break dwarf spheroidal galaxies (with R. Thiru Senthil) Arxiv: 1906.08951.

\item Does the Wolfenstein form work for the leptronic mixing matrix? arxiv: 1907.08380.

\item Superstrings: Another view (with San Fu Tuan) Physics Today, March 1987, p15.

\item Is quantum Mechanics for ever? Talk at the meeting on Philosphical Foundations of Quantum Theory, March 24-26, 1988, New Delhi (Proceeding, ed: Ranjit Nair, World Scientific).

\item Quantum interference effects in a five-dimensional Kaluza-Klein Theory (with R. Parthasarathy and R. Vasudevan), Classical and Quantum Gravity 3, 625 (1986).
\end{enumerate}

\newpage
