\makeatletter
\def\@makeschapterhead#1{%
  \vspace*{50\p@}%
  {\parindent \z@ \raggedright
    \normalfont
    \interlinepenalty\@M
    \LARGE \bfseries  #1\par\nobreak
    \vskip 40\p@
  }}
\makeatother

\chapter*{Acknowledgements}\label{ack1}

\lhead[\small\thepage\quad Reclaiming Rāmāyaṇa : Disentangling the Discourses]{}
\rhead[]{Acknowledgements\quad\small\thepage}

To Sri Rajiv Malhotra I owe an immense debt of gratitude. It was he who first drew my attention --- through his books --- to the existence of the Western-academia-challenge that looms before us. It was at his conference that I learnt of the severity of the problem --- and the urgency of a rebuttal. It was he who first suggested that I write this monograph --- and graciously made place in his institution for me. It was he who channeled the financial support to complete it --- without him, and his institution --- Infinity Foundation India --- this monograph would truly be inconceivable.   

I am deeply indebted to Prof. K. S. Kannan who was equally --- if not more --- instrumental in the undertaking of this monograph. But for him, I would not have met Sri Rajiv Malhotra. But for him, I would not have been chosen to undertake this work. But for him, I would not have had the courage or confidence to take it up --- or do justice to it. His support has been unflagging --- I have depended on his learning in moments of confusion, on his ingenuity to counter that of Prof. Sheldon Pollock. A constant source of encouragement, he has, with utmost patience, discussed with me my ideas, offered constructive criticism and made countless suggestions for improvements throughout the months that I have worked on this monograph.  I gratefully acknowledge Ms. H. R. Meera’s support throughout this project. I have had endlessly enriching discussions with her; many --- if not most --- ideas have stemmed from them.  She has also expertly dealt with the final re-touching of the monograph, laboring hard to ensure editorial consistency --- I am very grateful for it. 

Each of the following friends and IFI team members have contributed in various ways, and I regret I do not have the space to explain how valuably: Śatāvadhāni Dr. R. Ganesh, Alok Mishra, Shashi Kiran B.N., Shalini Puthiyedam, Vijaya Vishwanathan, and Sudarshan T. N. A generous funding from Vellayan Chettiar Trust has facilitated this work --- I am deeply grateful for it. 


\bigskip
\hfill {\bf Manjushree Hegde}

