\chapter[The Dark Universe]{The Dark Universe}\label{chap14}

\Authorline{}
\addtocontents{toc}{\protect\contentsline{section}{{\sl }\smallskip}{}}
\authinfo{}
\lhead[\small\thepage]{\small\thechapter. }



About 95 percent of the Universe is dark. Whatever we see
(humans, animals, plants, Sun, stars, galaxies) is only
5 percent of the Universe. All of these are made of atoms
and atoms are made of only three kinds of particles electrons,
protons and netrons. We now know much about these but that is
only 5 percent of everything.

According to astrophysicists and cosmologists, 70 percent of
the Universe is Dark Energy and 25 percent is Dark Matter.
Research is going on about this 70 + 25 = 95 percent Dark
Universe, but no definitive conclusions have been reached.
We shall first discuss the 25 percent Dark matter before we
go to the stranger Dark Energy.

\section*{Dark Matter}

Dark Matter w3as discovered by astronomers many years ago.
Both visible matter and Dark Matter obey Newton;s Law of
Gravitation. Using this, they discovered that Dark Matter
existed throughout the Universe.

There are many galaxies and galaxy clusters in the Universe.
Since the galaxies in a cluster attract each other, all
the galaxies are moving around the centre of the cluster.
This motion follows Newton's law of gravitation and hence
from this motion the mass of the whole cluster can be
calculated. This mass can also be calculated by adding all
the masses of the galaxies. The mass calculated from the
velocities was many times larger than the masses obtained
by adding the masses of the galaxies. So it was concluded
that invisible or Dark Matter existed so that the two
calculations would agree.

Next let us consider the motion of stars in the galaxy.
All the stars are moving around the centre of the galaxy
because of the gravitational attraction. That is why
many galaxies appear to be rotating. From the velocities
of the stars and Newton's law the total mass of the
galaxy was calculated. By adding the masses of all the
stars also the total mass of the galaxy could be obtained.
Again the mass calculated from the stellar velocities
was many times larger than the mass obtained by adding
the stellar masses. So again the existence of Dark Matter
was inferred.

Lord Kelvin was the first to guess the existence of Dark
Matter. He inferred this in 1884 from his studies of
stellar velocities. In 1906 Henri Poincare described
Kelvin's results in his essay and named it as Dark Matter.
Fritz Zwicky and Vera Rubin were the astronomers who
found out the precise details about Dark Matter. In
1938 Zwicky found the Dark Matter from the velocities
of galaxies in galactic clusters. Rubin found it from
the stellar velocities in galaxies.

After that, the existence of Dark Matter was confirmed
from many astronomical and cosmological researches. We
will mention those briefly. They are gravitational lensing,
Cosmic Microwave Background Radiation, cosmologicalevolution and structure formation in the Universe.
In the sixteenth and seventeenth centuries the existence
of atmosphere around the Earth was discovered. Everybody
knows the importance of the atmosphere consisting Oxygen
and Nitrogen. Dark Matter is more important than atmosphere
since the latter exists only for a height of 30 or 40
kilometers while DArk Matter exists all over the Universe.
Although astronomers have discovered the existence of Dark
Matter through gravitaional force, its nature can be discovered
only by physicists.

Such research in physics has been going on for more than
20 years. The basis of this research is the assumption that
Dark Matter is made of particles not yet discovered. Although
these particles interact much with the known particles
such as proton, electron etc., if they interact weakly,
many kinds of experiments are possible. We shall describe
this research briefly.

\section{Direct Interaction}

Since Dark Matter Particles (DMP) are everywhere, they may
sometime collide with nuclei of ordinary matter. These
nuclei will move and ionize the surrounding matter and hence
can be detected by particle detectors. Many such experiments
are being conducted for the past 20 years, but so far definitive
positive results have not been obtained.

\section{Indirect Detection}

A DMP may collide with another DMP and particles such as proton,
electron and photon may be produced. Such collisions will occur
only in those locations where DMP's are abundant. Since it is
believed that Dark Matter abounds at the galactic centres and
the cores of the Sun and other stars, experimenters are searching
for energetic particles coming from these directions. But these
searches also have not succeeded so far.

\section{Production by particle accelerators}

When two protons collide with high energy, many particles are
produced. In 2012, Higgs boson was discovered in such collisions
at the Large Hadron Collider. So DMP also may be produced in such
collisions. With this expectation, many experiments are being
conducted at the collider, but without any success so far.

Since experiments conducted for many years all over the world have
not succeeded, the question arises:

Does Dark Matter really exist?

As already mentioned, sofar, Dark Matter has been inferred only
from its gravitational interaction. All the Dark Matter experiments
assume that dark Matter interacts weakly with ordinary matter. If
this assuption is wrong, above experiments will never catch Dark
Matter. In other words if Dark Matter has only gravitational
interaction and does not have electromagnetic, weak, strong
or any other interaction with ordinary matter, we will never
know the true nature of Dark Matter. This is one way. There
are three other ways.

The second way is for some reason the density of Dark Matterin our region is very small, in which case the direct detection
experiment done on the Earth cannot catch it. But the other two
experiments can detect it.

The third possibility is Dark Matter may not be made of particles.
Dark Matter may be the primordial black holes. Just as all matter
was produced in the Big Bang, primordial black holes also might
have been produced and that may be the Dark Matter.

The fourth possibility is to change Newton's law of gravitation
(and Einstein's gravitational field equations) a little. Then there
will be no need for Dark Matter since Newton's law was used in getting
Dark Matter. If this possibility turns out to be true, it will be
a big revolution in Physics and Astronomy!
Among the four possilities above, which is correct? Or is there
any other way? Continuing research only can answer this. So
at present we cannot say anything definite about Dark Matter.

\section*{Dark Energy}

Energy is Mass, Mass is Energy. Eistein's equation $E = Mc^{2}$
declares this. In spite of this, we talk about Dark Matter and
Dark Energy as two separate things. Why? Reason is: Dark Matter
has mass and other properties like visible matter, but Dark
Energy is not matter at all. It is pure energy!
To understand Dark Energy, mathematics and Einstein's General
Theory of Relativity are neccssary. So it is difficult to explain
it here.

Nevertheless, some facts have to be mentioned. From Einstein's
gravitational equations, it is possible to calculate the evolution
of the Universe. How the Universe is expanding from the time of
its birth in the Big Bang? This can be explained from Einstein
equations. That is what cosmologists also found.

But, in 1999, a major discovery was made, namely the speed of the
expansion of the Universe is increasing. This discovery was made
by studies on a large number of supernovae. So one had to include
Dark Energy in Eistein's equations. That is required to explain
the accelerating Unverse. This Dark Energy can also be called the
Energy of the Vaccuum.

Another consequence of Eistein's General Relativity is the presence
of Dark Pressure along with Dark Energy. This Dark Prssure is not
like the pressure due to gases with which we are familiar. It is
a negative pressure. From the discovery of the acceleration of the
Universe, Dark Energy and Dark Pressure have been inferred.

If there were no Dark Energy, there would have been only one constant,
namely Newton's constant of Gravitation G in Einstein's equation. Presence
of Dark Energy neccssitates the addition of another constant, Lambda,
called Cosmological Constant in Eistein's equation.

This Cosmological Constant has an interesting history. After Einstein
discovered his gravitational equations,he used them to study the
evolution of the Universe. In that period of time, Universe was
thought to be in a steady state. But Einstein's equations contradicted
the steady state. So Einstein included the cosmological constant Lambda
in his equations to stabilise the Universe. He did not like this, since
he had been very happy that his original equations did not have any new
constant apart from G, but still they led to so many new consequences.The addition of Lambda destroyed that feeling.
But after Einstein's addition of Lambda, two major developments took
place in Cosmology. In 1924 the Russian mathematician Friedman found an exact
solution of the original Einstein equations and this solution described
an expanding Universe. Second, in 1929, astronomer Edwin Hubble discovered
that all the galaxies that he studied through the big 100 inch telescope in
California appeared to be running away from us. If the Universe were
expanding, this would happen. Thus Friedman through theory and Hubble
through observation discovered the expansion of the Universe.

Immediately Einstein removed the Lambda that he added in his equations
against his will. He called the addition of Lambda as his greatest
blunder. For, if he had not done that, he could have predicted the
expansion of the Universe from his original equations.
During that period of time, cosmology had not advanced much. After
that, cosmological research progressed during the 80 years and
reached the stage when the acceleration of the Universe could be
discovered. So Einstein's Lambda was brought back. As we mentioned
already, it refers to the vaccuum energy.

Nevertheless, we cannot be sure that this is the real reason for
the acceleration. Instead of Lambda, an all-pervasive field may
be the cause. This field may be the vaccuum energy. There may
be other possibilities. Reserach is going on. Future only give
the answer.

We understand only 5 percent of the Universe. The major part of
95 percent is dark. Although active research is being conducted
all over the world to find the true nature of this 95 percent,
the truth is still not known. So we see how strange our Universe
is. It is full of unsolved mysteries.
