
\thispagestyle{empty}
~\phantom{a}
\vfill

\begin{center}
{\Huge\bf pu. ti. na}\\[15pt]
{\Huge\bf avara}\\[15pt]
{\Huge\bf samagarx anuvAdita leVKanagaLu}
\end{center}

\vfill
\eject

\thispagestyle{empty}
~\phantom{a}
\vfill

\vfill
\eject

\thispagestyle{empty}
~\phantom{a}

\begin{center}
{\Huge\bf pu. ti. na. avara}\\[5pt]
{\Huge\bf samagarx anuvAdita leVKanagaLu}
\end{center}

\vfill

\begin{center}
{\bf saMpAdaka}\\[5pt]
{\Large\bf DA. caMdarxsheVKara enf. beTaTxhaLiLx}
\end{center}

\vfill

\begin{center}
{\large\bf DA. pu.~ti.~na.~ TarxsfTx (ri).}\\[2pt]
na, 20, saMkalapx, puSapxgirinagara, hosakerehaLiLx,\\[2pt]
banashaMkari 3neV haMta, beMgaLUru~-~560085
\end{center}

\eject

\thispagestyle{empty}

\noindent
{\eng\fontsize{8}{8pt}\selectfont\bfseries PU.~THI.~NA. AVARA SAMAGRA ANUVADITA LEKHANAGALU}
\bigskip

\noindent
{\eng\fontsize{10}{12}\selectfont An arthology of complete trarslated articals of Dr.~Pu.~Thi. Narasimhachar, Edited by DR. Chandrashekara N. Bettahalli, Published by PU. THI. NA. Trust, No.~30, Sankalpa, Pushvagirinagara, Hosakerehalli, Banashankare, 3rd Stage, Bengaluru - 560085}
\vskip 1.2cm

\noindent
{\eng\fontsize{10}{12}\selectfont First Edition : 2017}
\vskip 1.2cm

\noindent
{\eng\fontsize{10}{12}\selectfont 
Pages: \pageref{prelims:end} + \pageref{book:end}\\[3pt]
Paper: N. S. Maplitho 70 GSM\\
Size: Crown 1/8\\[3pt]
\copyright~ Author\\[30pt]
\rupee \\[30pt]
{\eng\bfseries Publisher:}\\ 
PU. THI. NA. Trust\\
No. 30, Sankalpa.\\
Pushvagirinagara\\
Hosakerehalli\\
Banashankara, 3rd Stage\\
Bengaluru - 560085\\[15pt]
{\eng\bfseries Typeset By:} Sriranga Digital Software Technologies Private Limited\\
Srirangapatna\\[15pt]
{\eng\bfseries Printer:}}

\newpage

\thispagestyle{empty}
~\phantom{a}

\vfill

\begin{minipage}[c]{8cm}
\begin{center}
pu. ti. na. avara Apatx senxVhitarU,\\[3pt]
parxsidadhx sAhitigaLU,\\[3pt]
Ada~ {\large\bf DA. parxBushaMkara}~ avarige\\[3pt]
pirxVtipUravxkavAgi
\end{center}

\bigskip

\begin{flushright}
saMpAdaka\hspace{1.2cm}\,\\[3pt]
{\bf DA. caMdarxsheVKara enf. beTaTxhaLiLx}
\end{flushright}
\end{minipage}

\vfill\eject

\thispagestyle{empty}
~\phantom{a}

\begin{center}
{\Large\bf pu. ti. na. TarxsfTx (ri)}\\[3pt]
naM. {\rm 30,} saMkalapx, puSapxgirinagara, hosakerehaLiLx\\
banashaMkari {\rm3}neV haMta, beMgaLUru - {\rm 560085}\\[10pt]
{\large\bf kAyaRkAri maMDaLi}\\[7pt]
mahApoVSakaru\\[3pt]
rASaTxrXkavi ji. esf. shivarudarxpapx\\[15pt]
adhayxkaSxru\\[3pt]
DA. ecf. esf. veMkaTeVshamUtiR\\[15pt]
kAyaRdashiR\\
shirxV KAdirx esf. acuyxtanf\\[15pt]
sadasayxrugaLu\\[5pt]
DA. parxBushaMkara\\[3pt]
shirxV veY. ke. mududxkaqSaNx\\[3pt]
DA. shirxVrAmaBaTaTx\\[3pt]
DA. bi. vi. rAjArAmf\\[3pt]
DA. Di, bAlakaqSaNx\\[3pt]
shirxV enf. esf. shirxVnAthf\\[15pt]
padanimitatx sadasayxru\\[3pt]
Ayukatxru, kananxDa matutx saMsakxti ilAKe\\[3pt]
jilAlxdhikArigaLu, maMDayx
\end{center}



